### 2024-03-08 models/negative_bin_model.Rmd
Summary information
created tables and graphs showing the leading causes of death for overall population and by race

---
title: "negative_bin_model.Rmd"
format: html
---

```{r}
# load libraries
library(MASS)
library(dplyr)

# load files 
file_path <- "../scripts/death_count_by_Cause.csv"
death_count_by_cause <- read.csv(file_path)

death_count_by_cause$Leading.Cause <- as.factor(death_count_by_cause$Leading.Cause)

# Attempt to increase the iteration limit to allow convergence
nb_model <- glm.nb(Total_Deaths ~ Leading.Cause, 
                   data = death_count_by_cause, 
                   control = glm.control(maxit = 1000)) # Increase max iterations to 1000

# Check the summary
summary(nb_model)

```

### 2024-03-08 scripts/leading_cause_of_death.Rmd
Summary information
created tables and graphs showing the leading causes of death for overall population and by race

---
title: "leading_cause_of_death_overall"
format: html
---

```{r}
# libraries
library(dplyr)
library(knitr)
library(ggplot2)
library(forcats)

# read in the dataset

file_path <- "../data/analysis_data/clean_nyc_data.csv"
nyc_data <- read.csv(file_path)

```

```{r}
# create table showing count of deaths by cause (overall)
death_count_by_cause <- nyc_data %>%
  group_by(Leading.Cause) %>%
  summarise(Total_Deaths = sum(as.numeric(Deaths), na.rm = TRUE)) %>%
  arrange(desc(Total_Deaths))
knitr::kable(death_count_by_cause, caption = "Count of Deaths by Leading Cause")
write.csv(death_count_by_cause, "death_count_by_cause.csv", row.names = FALSE)
```

```{r}
# create table showing top 4 leading causes of death by race
top_causes_by_race <- nyc_data %>%
  group_by(Race.Ethnicity, Leading.Cause) %>%
  summarise(Total_Deaths = sum(as.numeric(Deaths), na.rm = TRUE)) %>%
  arrange(Race.Ethnicity, desc(Total_Deaths)) %>%
  mutate(rank = row_number()) %>%
  filter(rank <= 4) %>%
  ungroup()

table_top_causes_by_race <- top_causes_by_race %>%
  select(-rank) %>%
  arrange(Race.Ethnicity, desc(Total_Deaths))

knitr::kable(table_top_causes_by_race, caption = "Top 4 Leading Causes of Death by Race")

# create bar graph to display leading causes of death by race
gg <- ggplot(top_causes_by_race, aes(x = Leading.Cause, y = Total_Deaths)) +
  geom_col(aes(fill = Leading.Cause)) + 
  facet_wrap(~Race.Ethnicity, scales = "free_y") + 
  scale_fill_brewer(palette = "Paired") + 
  theme_minimal() + # Minimal theme
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        strip.text.x = element_text(size = 8), 
        legend.position = "none") + #
  labs(x = "Top 4 Causes of Death", y = "Total Deaths")

print(gg)
ggsave("top_causes_by_race.png", plot = gg, width = 12, height = 8)

```

### 2024-03-09 scripts/leading_cause_of_death.Rmd
model fixes
Updated the negative binomial regression models and cleaned up the scripts folder

@@
 ```{r}
-# create table showing top 4 leading causes of death by race
+# create table showing top 5 leading causes of death by race
 top_causes_by_race <- nyc_data %>%
@@
   mutate(rank = row_number()) %>%
-  filter(rank <= 4) %>%
+  filter(rank <= 5) %>%
   ungroup()
@@
 
-knitr::kable(table_top_causes_by_race, caption = "Top 4 Leading Causes of Death by Race")
+knitr::kable(table_top_causes_by_race, caption = "Top 5 Leading Causes of Death by Race")
 
@@
   scale_fill_brewer(palette = "Paired") + 
-  theme_minimal() + # Minimal theme
-  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
-        strip.text.x = element_text(size = 8), 
-        legend.position = "none") + #
-  labs(x = "Top 4 Causes of Death", y = "Total Deaths")
+  theme_minimal() + 
+  theme(axis.text.x = element_text(angle = 40, hjust = 1), 
+        strip.text.x = element_text(size = 6), 
+        legend.position = "none") + 
+  labs(x = "Top 5 Causes of Death", y = "Total Deaths")
 

### 2024-03-09 scripts/model_asianpacific.Rmd
model fixes
Updated the negative binomial regression models and cleaned up the scripts folder

---
title: "model_asianpacific"
format: html
---

```{r}
# load libraries
library(dplyr)
library(readr)
library(ggplot2)
library(forcats)
library(rstanarm)
library(knitr)
library(kableExtra)
```

```{r}
# load the dataset 
file_path <- "../data/analysis_data/clean_nyc_data.csv"
nyc_data <- read.csv(file_path)

```

```{r}
# Convert Deaths to numeric if it's not already
nyc_data$Deaths <- as.numeric(as.character(nyc_data$Deaths))

# Filter for the year 2014 and the specific race
race_name <- "Asian and Pacific Islander"

nyc_data_2014_race <- nyc_data %>%
  filter(Year == 2014, Race.Ethnicity == race_name)

# Identify the top 5 causes of death for the selected race
top_causes_by_race <- nyc_data_2014_race %>%
  group_by(Leading.Cause) %>%
  summarise(Total_Deaths = sum(as.numeric(Deaths), na.rm = TRUE), .groups = 'drop') %>%
  arrange(desc(Total_Deaths)) %>%
  slice_max(Total_Deaths, n = 5)

# Ensure data for regression includes only the top causes
nyc_data_for_regression <- nyc_data_2014_race %>%
  inner_join(top_causes_by_race, by = "Leading.Cause")

```

```{r}
# negative binomial regression
neg_binom_model <- stan_glm(
  Total_Deaths ~ Leading.Cause,
  data = nyc_data_for_regression,
  family = neg_binomial_2(link = "log"),
  seed = 853
)

# Output the model summary
model_summary <- summary(neg_binom_model)

# Display the model summary table
kable(model_summary, caption = paste("Negative Binomial Regression Summary for", race_name)) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### 2024-03-12 models/model_overall_causes.Rmd
cleaned up repo + sketches
Cleaned up repository to remove additional/unwanted files. Renamed model. Added sketches. Wrote the README.

---
title: "negative_bin_model.Rmd"
format: html
---

```{r}
# load libraries
library(dplyr)
library(readr)
library(ggplot2)
library(forcats)
library(rstanarm)
library(knitr)
library(kableExtra)
```

```{r}
# load the dataset 
file_path <- "../data/analysis_data/clean_nyc_data.csv"
nyc_data <- read.csv(file_path)

```

```{r}
# Filter for the year 2014
nyc_data$Deaths <- as.numeric(as.character(nyc_data$Deaths))
nyc_data_2014 <- nyc_data %>%
  filter(Year == 2014)

# Identify top 5 causes of death overall
top_causes_overall <- nyc_data_2014 %>%
  group_by(Leading.Cause) %>%
  summarise(Total_Deaths = sum(Deaths, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(desc(Total_Deaths)) %>%
  slice_max(order_by = Total_Deaths, n = 5)

# Filter data for top 5 causes
nyc_data_2014_top_causes_with_totals <- nyc_data_2014 %>%
  inner_join(top_causes_overall, by = "Leading.Cause")

```

```{r}
# Negative binomial regression for overall top 5 causes
neg_binom_model_overall <- stan_glm(
  Total_Deaths ~ Leading.Cause,
  data = nyc_data_2014_top_causes_with_totals,
  family = neg_binomial_2(link = "log"),
  seed = 853
)
```

```{r}
# Summary model
summary(neg_binom_model_overall) 
```

### 2024-03-12 scripts/leading_cause_of_death.Rmd
cleaned up repo + sketches
Cleaned up repository to remove additional/unwanted files. Renamed model. Added sketches. Wrote the README.

@@
 knitr::kable(death_count_by_cause, caption = "Count of Deaths by Leading Cause")
-write.csv(death_count_by_cause, "death_count_by_cause.csv", row.names = FALSE)
+
 ```

### 2024-03-14 paper/paper.qmd
Update Paper

---
title: "Beyond the Numbers: Dissecting New York City's Leading Causes of Death Across Demographics"
subtitle: "A demographics-based analysis of morbditiy in the Big Apple"
author: 
  - Rayan Awad Alim
  - Maria Mangru
  - MD Mubtasim-Fuad 
thanks: "Code and data are available at: https://github.com/RayanAlim/Analysis-of-Morbidity-in-New-York"
date: today
date-format: long
abstract: "We analysed the leading causes of death in New York City, segmented by sex and ethnicity, from 2007 to 2014. Utilizing data derived from NYC death certificates, our research offers insights into public health trends, and the importance of understanding how these causes of death vary across race."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
```

# Introduction

You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.

The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

DATA 
This analysis utilizes data from the New York City Department of Health and Mental Hygiene, specifically provided by the Bureau of Vital Statistics {@NYCData}. The dataset encompasses records of deaths in New York City since 2007, detailing the leading causes of death categorized by sex and ethnicity. Each entry in this dataset is derived from NYC death certificates which are the official documentation of every death occurring within the city's jurisdiction.

Dataset Context and Broader Implications
The investigation into the leading causes of death within New York City holds significant public health importance. It allows for a nuanced understanding of mortality trends, guiding health policy, and intervention strategies tailored to specific demographics and causes. By examining mortality across different sexes and ethnic groups, this analysis contributes to identifying health disparities and targeting efforts to address them.

Variables and Data Examination
The dataset comprises several key variables, including:
Year: The year of the recorded death.
Leading Cause: The primary cause of death as categorized by ICD-10 codes.
Sex: The sex of the deceased (Male, Female, or Gender Non-Conforming).
Race Ethnicity: The self-reported ethnicity of the deceased, including categories such as Hispanic, White Non-Hispanic, Black Non-Hispanic, Asian and Pacific Islander, and Others.
Deaths: The total number of deaths attributed to the leading cause.
Death Rate and Age-Adjusted Death Rate: Rates per 100,000 population, providing standardized comparisons across different population sizes and age structures.

Alternative Datasets and Justification
While other mortality databases exist, such as the CDC's National Center for Health Statistics, the chosen dataset offers a detailed and localized perspective specific to New York City. This specificity provides a more precise tool for understanding and addressing urban health dynamics, making it more relevant for city-specific policy-making and health intervention strategies.

Data Cleaning and Variable Construction
Preliminary data cleaning focused on addressing missing values, ensuring consistency in categorical variables, and suppressing unreliable rates as mentioned. New variables, such as aggregated categories for cause of death or demographic groupings, were constructed to facilitate analysis. For instance, causes of death were grouped into broader categories (e.g., cardiovascular diseases, cancers) to examine trends at a macro level.

Summary Statistics and Relationships
Initial exploratory data analysis revealed key trends and disparities in mortality rates across different populations within New York City. We provide summary statistics, including mean death rates and distributions of deaths by cause ethnicity, to provide insights into the health landscape of the city. For example, preliminary findings indicate significant differences in heart disease mortality between ethnic groups, warranting further investigation.

Measurement and Methodological Notes
The measurement of mortality and its causes relies on the accurate classification of death certificates, adhering to ICD-10 standards. This classification ensures comparability with other datasets and robustness in identifying health trends. However, it's important to acknowledge potential limitations in cause-of-death reporting and classification, which may impact the analysis. To ensure privacy and reliability, rates based on small numbers (Relative Standard Error, RSE, > 30) and aggregate counts less than 5 have been suppressed. This suppression safeguards against the identification of individuals in rare categories and ensures statistical reliability.

# Model

### Model justification

# Results

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

\newpage

# References

### 2024-03-14 scripts/leading_cause_of_death.Rmd
Created simulated data and tests
Created tests for simulated data and for leading cause of death.

@@
 library(forcats)
+library(testthat)
 
@@
 ```
+
+```{r}
+# Tests
+
+# Data set has been loaded correctly
+test_that("Dataset has been loaded", {
+  expect_true(exists("nyc_data"))
+  expect_is(nyc_data, "data.frame")
+})
+
+# 'death_count_by_cause' has the correct columns
+test_that("'death_count_by_cause' has correct columns", {
+  expected_cols <- c("Leading.Cause", "Total_Deaths")
+  expect_equal(colnames(death_count_by_cause), expected_cols)
+})
+
+# 'death_count_by_cause' table has correct number of rows
+test_that("'death_count_by_cause' has correct number of rows", {
+  expect_true(nrow(death_count_by_cause) > 0)
+})
+
+```

### 2024-03-14 scripts/simulated_dataset.Rmd
Created simulated data and tests
Created tests for simulated data and for leading cause of death.

---
title: "simulated dataset"
output: html_document
---

```{r}
# Load libraries
library(dplyr)
library(tidyr)
library(stringr)
library(testthat)

set.seed(853) # Set a seed for reproducibility

```

```{r}
# Number of rows to simulate
n <- 500

# Leading causes of death and their ICD-10 codes
leading_causes <- c(
  "Human immunodeficiency virus (HIV) disease B20-B24",
  "Malignant neoplasms C00-C97",
  "Diabetes mellitus E10-E14",
  "Alzheimer’s disease G30",
  "Diseases of heart I00-I09,I11,I13,I20-I51",
  "Essential hypertension and hypertensive renal disease I10, I12",
  "Cerebrovascular diseases I60-I69",
  "Atherosclerosis I70",
  "Other diseases of circulatory system I71-I78",
  "Influenza and pneumonia J10-J18",
  "Chronic lower respiratory diseases J40-J47",
  "Chronic liver disease and cirrhosis K70,K73-K74",
  "Nephritis, nephrotic syndrome and nephrosis N00-N07,N17-N19,N25-N27",
  "Certain conditions originating in the perinatal period P00-P96",
  "Congenital malformations, deformations & chromosomal abnormalities Q00-Q99",
  "Sudden infant death syndrome R95",
  "Symptoms, signs and abnormal clinical and laboratory findings, not elsewhere classified R00-R94,R96-R99",
  "All other diseases (Residual)",
  "Motor vehicle accidents V02-V04,V09.0,V09.2,V12-V14,V19.0-V19.2,V19.4-V19.6,V20-V79,V80.3-V80.5,V81.0-V81.1,V82.0-V82.1,V83-V86,V87.0-V87.8,V88.0-V88.8,V89.0,V89.2",
  "All other and unspecified accidents and adverse effects V01,V05-V06,V09.1,V09.3-V09.9,V10-V11,V15-V18,V19.3,V19.8-V19.9,V80.0-V80.2,V80.6-V80.9,V81.2-V81.9,V82.2-V82.9,V87.9,V88.9,V89.1,V89.3,V89.9,V90-X59,Y40-Y86,Y88",
  "Intentional self-harm (suicide) X60-X84,Y87.0",
  "Assault (homicide) X85-Y09,Y87.1",
  "All other external causes Y10-Y36,Y87.2,Y89"
)

# Generate the data
simulated_data <- tibble(
  Year = sample(2007:2014, n, replace = TRUE),
  Leading_Cause = sample(leading_causes, n, replace = TRUE),
  Sex = sample(c("M", "F"), n, replace = TRUE),
  Race_Ethnicity = sample(c("White Non-Hispanic", "Hispanic", "Black Non-Hispanic", 
                             "Asian and Pacific Islander", "Not Stated/Unknown", "Other Race/Ethnicity"), n, replace = TRUE),
  Deaths = sample(1:1500, n, replace = TRUE),
  Death_Rate = runif(n, 5, 30),
  Age_Adjusted_Death_Rate = runif(n, 5, 30)
)

```

```{r}
# Tests for generated data

# Ensure the simulated dataset has properties similar to the original dataset
test_that("Data has correct types", {
  expect_is(simulated_data$Year, "integer")
  expect_is(simulated_data$Leading_Cause, "character")
  expect_is(simulated_data$Sex, "character")
  expect_is(simulated_data$Race_Ethnicity, "character")
  expect_is(simulated_data$Deaths, "integer")
  expect_is(simulated_data$Death_Rate, "numeric")
  expect_is(simulated_data$Age_Adjusted_Death_Rate, "numeric")
})

# Ensure 'Year' column contains only the years in the specified range
test_that("Year column contains appropriate years", {
  expect_true(all(simulated_data$Year >= 2007 & simulated_data$Year <= 2014))
})

# Ensure no duplicated rows exists
test_that("There are no duplicate rows", {
  expect_equal(nrow(simulated_data), nrow(distinct(simulated_data)))
})

```

### 2024-03-14 paper/paper.qmd
Update paper.qmd to include datasheet in appendix

@@
 bibliography: references.bib
+editor: 
+  markdown: 
+    wrap: sentence
 ---
@@
 
-You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.
+You can and should cross-reference sections and sub-sections.
+We use @citeR and @rohan.
 
-The remainder of this paper is structured as follows. @sec-data....
+The remainder of this paper is structured as follows.
+@sec-data....
 
@@
 
-DATA 
-This analysis utilizes data from the New York City Department of Health and Mental Hygiene, specifically provided by the Bureau of Vital Statistics {@NYCData}.
+DATA This analysis utilizes data from the New York City Department of Health and Mental Hygiene, specifically provided by the Bureau of Vital Statistics {@NYCData}.
 The dataset encompasses records of deaths in New York City since 2007, detailing the leading causes of death categorized by sex and ethnicity.
@@
 
-Dataset Context and Broader Implications
-The investigation into the leading causes of death within New York City holds significant public health importance.
+Dataset Context and Broader Implications The investigation into the leading causes of death within New York City holds significant public health importance.
 It allows for a nuanced understanding of mortality trends, guiding health policy, and intervention strategies tailored to specific demographics and causes.
@@
 
-Variables and Data Examination
-The dataset comprises several key variables, including:
-Year: The year of the recorded death.
+Variables and Data Examination The dataset comprises several key variables, including: Year: The year of the recorded death.
 Leading Cause: The primary cause of death as categorized by ICD-10 codes.
@@
 
-Alternative Datasets and Justification
-While other mortality databases exist, such as the CDC's National Center for Health Statistics, the chosen dataset offers a detailed and localized perspective specific to New York City.
+Alternative Datasets and Justification While other mortality databases exist, such as the CDC's National Center for Health Statistics, the chosen dataset offers a detailed and localized perspective specific to New York City.
 This specificity provides a more precise tool for understanding and addressing urban health dynamics, making it more relevant for city-specific policy-making and health intervention strategies.
 
-Data Cleaning and Variable Construction
-Preliminary data cleaning focused on addressing missing values, ensuring consistency in categorical variables, and suppressing unreliable rates as mentioned.
+Data Cleaning and Variable Construction Preliminary data cleaning focused on addressing missing values, ensuring consistency in categorical variables, and suppressing unreliable rates as mentioned.
 New variables, such as aggregated categories for cause of death or demographic groupings, were constructed to facilitate analysis.
@@
 
-Summary Statistics and Relationships
-Initial exploratory data analysis revealed key trends and disparities in mortality rates across different populations within New York City.
+Summary Statistics and Relationships Initial exploratory data analysis revealed key trends and disparities in mortality rates across different populations within New York City.
 We provide summary statistics, including mean death rates and distributions of deaths by cause ethnicity, to provide insights into the health landscape of the city.
@@
 
-Measurement and Methodological Notes
-The measurement of mortality and its causes relies on the accurate classification of death certificates, adhering to ICD-10 standards.
+Measurement and Methodological Notes The measurement of mortality and its causes relies on the accurate classification of death certificates, adhering to ICD-10 standards.
 This classification ensures comparability with other datasets and robustness in identifying health trends.
 However, it's important to acknowledge potential limitations in cause-of-death reporting and classification, which may impact the analysis.
-To ensure privacy and reliability, rates based on small numbers (Relative Standard Error, RSE, > 30) and aggregate counts less than 5 have been suppressed.
+To ensure privacy and reliability, rates based on small numbers (Relative Standard Error, RSE, \> 30) and aggregate counts less than 5 have been suppressed.
 This suppression safeguards against the identification of individuals in rare categories and ensures statistical reliability.
@@
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+If my paper were 10 pages, then should be be at least 2.5 pages.
+The discussion is a chance to show off what you know and what you learnt from all this.
 
@@
 
-# Appendix {-}
+# Appendix {.unnumbered}
 
@@
 
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
+In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check.
+This shows...
+
+# Datasheet
+
+## Motivation
+
+-   For what purpose was the dataset created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.
+    -   This dataset was compiled to facilitate the analysis of mortality patterns in New York City, providing insights into the leading causes of death across the general population and by race.
+-   Who created the dataset (for example, which team, research group) and on behalf of which entity (for example, company, institution, organization)?
+    -   This dataset was created by a research team at the New York City Department of Health and Mental Hygiene.
+-   Who funded the creation of the dataset? If there is an associated grant, please provide the name of the grantor and the grant name and number.
+    -   Funding was provided as part of the budget of the New York City Department of Health and Mental Hygiene.
+-   Any other comments?
+    -   No.
+
+## Composition
+
+-   What do the instances that comprise the dataset represent (for example, documents, photos, people, countries)?
+    Are there multiple types of instances (for example, movies, users, and ratings; people and interactions between them; nodes and edges)?
+    Please provide a description.
+
+    -   Each row in the dataset represents an aggregate count of deaths for a specific leading cause, categorized by year, sex, and race/ethnicity.
+
+-   How many instances are there in total (of each type, if appropriate)?
+
+    -   Over 1000 instances.
+
+-   Does the dataset contain all possible instances or is it a sample (not necessarily random) of instances from a larger set?
+    If the dataset is a sample, then what is the larger set?
+    Is the sample representative of the larger set (for example, geographic coverage)?
+    If so, please describe how this representativeness was validated/verified.
+    If it is not representative of the larger set, please describe why not (for example, to cover a more diverse range of instances, because instances were withheld or unavailable).
+
+    -   This dataset contains all recorded instances of death in New York City between 2007 - 2014.
+
+-   What data does each instance consist of?
+    “Raw” data (for example, unprocessed text or images) or features?
+    In either case, please provide a description.
+    Each instance consists of counts of death including demographic data (race and sex) and year.
+    Is there a label or target associated with each instance?
+    If so, please provide a description.
+
+    -   No.
+
+-   Is any information missing from individual instances?
+    If so, please provide a description, explaining why this information is missing (for example, because it was unavailable).
+    This does not include intentionally removed information, but might include, for example, redacted text.
+
+    -   Some instances may have missing data where the information was not recorded or reported.
+
+-   Are relationships between individual instances made explicit (for example, users’ movie ratings, social network links)?
+    If so, please describe how these relationships are made explicit.
+
+    -   Yes through demographic categorization (race and sex) and alignment.
+
+-   Are there recommended data splits (for example, training, development/validation, testing)?
+    If so, please provide a description of these splits, explaining the rationale behind them.
+
+    -   No.
+
+-   Are there any errors, sources of noise, or redundancies in the dataset?
+    If so, please provide a description.
+
+    -   Concerns arise given the potential for inaccuracies in the reporting and classification of death causes.
+
+-   Is the dataset self-contained, or does it link to or otherwise rely on external resources (for example, websites, tweets, other datasets)?
+    If it links to or relies on external resources, a) are there guarantees that they will exist, and remain constant, over time; b) are there official archival versions of the complete dataset (that is, including the external resources as they existed at the time the dataset was created); c) are there any restrictions (for example, licenses, fees) associated with any of the external resources that might apply to a dataset consumer?
+    Please provide descriptions of all external resources and any restrictions associated with them, as well as links or other access points, as appropriate.
+
+    -   Self-contained.
+
+-   Does the dataset contain data that might be considered confidential (for example, data that is protected by legal privilege or by doctor-patient confidentiality, data that includes the content of individuals’ non-public communications)?
+    If so, please provide a description.
+
+    -   Generated from public sources.
+
+-   Does the dataset contain data that, if viewed directly, might be offensive, insulting, threatening, or might otherwise cause anxiety?
+    If so, please describe why.
+
+    -   No.
+
+-   Does the dataset identify any sub-populations (for example, by age, gender)?
+    If so, please describe how these subpopulations are identified and provide a description of their respective distributions within the dataset.
+
+    -   Yes, sex is identified.
+
+-   Is it possible to identify individuals (that is, one or more natural persons), either directly or indirectly (that is, in combination with other data) from the dataset?
+    If so, please describe how.
+
+    -   No.
+
+-   Does the dataset contain data that might be considered sensitive in any way (for example, data that reveals race or ethnic origins, sexual orientations, religious beliefs, political opinions or union memberships, or locations; financial or health data; biometric or genetic data; forms of government identification, such as social security numbers; criminal history)?
+    If so, please provide a description.
+
+    -   This data does contain race origins and causes of death.
+
+-   Any other comments?
+
+    -   No.
+
+## Collection process
+
+-   How was the data associated with each instance acquired?
+Was the data directly observable (for example, raw text, movie ratings), reported by subjects (for example, survey responses), or indirectly inferred/derived from other data (for example, part-of-speech tags, model-based guesses for age or language)?
+If the data was reported by subjects or indirectly inferred/derived from other data, was the data validated/verified?
+If so, please describe how.
+    -   The data were gathered from death certificates and medical reports filed within New York City.
+-   What mechanisms or procedures were used to collect the data (for example, hardware apparatuses or sensors, manual human curation, software programs, software APIs)?
+How were these mechanisms or procedures validated?
+    -   Data was collected through official city and health databases. R scripts were used to clean and aggregate the data.
+-   If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?
+    -   Not a sample.
+-   Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?
+    -   New York City Department of Health and Mental Hygiene.
+-   Were any ethical review processes conducted (for example, by an institutional review board)?
+If so, please provide a description of these review processes, including the outcomes, as well as a link or other access point to any supporting documentation.
+    -   No.
+-   Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources (for example, websites)?
+    -   Collected from a third-party source.
+-   Were the individuals in question notified about the data collection?
+If so, please describe (or show with screenshots or other information) how notice was provided, and provide a link or other access point to, or otherwise reproduce, the exact language of the notification itself.
+    -   No.
+-   Any other comments?
+    -   No.
+
+## Preprocessing/cleaning/labeling
+
+-   Was any preprocessing/cleaning/labeling of the data done (for example, discretization or bucketing, tokenization, part-of-speech tagging, SIFT feature extraction, removal of instances, processing of missing values)?
+If so, please provide a description.
+If not, you may skip the remaining questions in this section.
+    -   Yes, data was cleaned to address missing or incomplete records.
+-   Was the “raw” data saved in addition to the preprocessed/cleaned/labeled data (for example, to support unanticipated future uses)?
+If so, please provide a link or other access point to the “raw” data.
+    -   Yes, both the “raw” and “cleaned” data were saved.
+-   Is the software that was used to preprocess/clean/label the data available? If so, please provide a link or other access point.
+    -   R was used.
+-   Any other comments?
+    -   No
+
+## Uses
+
+-   Has the dataset been used for any tasks already? If so, please provide a description.
+    -   Yes, the dataset has been used for annual health reports and studies on mortality trends within New York City.
+-    Is there a repository that links to any or all papers or systems that use the dataset? If so, please provide a link or other access point.
+    -   No.
+-   What (other) tasks could the dataset be used for?
+    -   The dataset could be used to predict future trends in public health.
+-   Is there anything about the composition of the dataset or the way it was collected and preprocessed/cleaned/labeled that might impact future uses?
+For example, is there anything that a dataset consumer might need to know to avoid uses that could result in unfair treatment of individuals or groups (for example, stereotyping, quality of service issues) or other risks or harms (for example, legal risks, financial harms)?
+If so, please provide a description.
+Is there anything a dataset consumer could do to mitigate these risks or harms?
+    -   No.
+-   Are there tasks for which the dataset should not be used? If so, please provide a description.
+    -   Discrimination of insurance rates/policies based on race.
+-   Any other comments?
+    -   No.
+
+## Distribution
+
+-   Will the dataset be distributed to third parties outside of the entity (for example, company, institution, organization) on behalf of which the dataset was created?
+If so, please provide a description.
+    -   The dataset is available through the NYC Open Data Profile.
+-   When will the dataset be distributed?
+    -   The dataset is uploaded and distributed annually through the NYC Open Data Profile.
+-   Have any third parties imposed IP-based or other restrictions on the data associated with the instances?
+If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms, as well as any fees associated with these restrictions.
+    -   None that are known.
+-   Do any export controls or other regulatory restrictions apply to the dataset or to individual instances If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any supporting documentation.
+    -   None that are known.
+-   Any other comments?
+    -   No.
+
+## Maintenance
+
+-   Who will be supporting/hosting/maintaining the dataset?
+    -   NYC Department of Health and Mental Hygiene.
+-   How can the owner/curator/manager of the dataset be contacted (for example, email address)?
+    -   A request can be sent to the NYC Open Data Portal team through their website (https://opendata.cityofnewyork.us/engage/)
+-   Will the dataset be updated (for example, to correct labeling errors, add new instances, delete instances)?
+If so, please describe how often, by whom, and how updates will be communicated to dataset consumers (for example, mailing list, GitHub)?
+    -   Yes, the dataset is updated annually.
+-   If others want to extend/augment/build on/contribute to the dataset, is there a mechanism for them to do so?
+If so, please provide a description.
+Will these contributions be validated/verified?
+If so, please describe how.
+If not, why not?
+Is there a process for communicating/distributing these contributions to dataset consumers?
+If so, please provide a description.
+    -   Contributions/suggestions can be made through the NYC Open Data portal.
+-   Any other comments?
+    -   No
 
@@
 
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
+@fig-stanareyouokay-1 is a trace plot.
+It shows...
+This suggests...
 
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
+@fig-stanareyouokay-2 is a Rhat plot.
+It shows...
+This suggests...
 

### 2024-03-15 scripts/02-cleaned_data.Rmd
changed to .r files and added tests

```{r}
library(tidyverse)
library(here)

#Reading the raw data

file_path <-
  here("data",
       "raw_data",
       "New_York_City_Leading_Causes_of_Death_20240306.csv")
raw_data <- read_csv(file_path)
```

```{r}
# Replacing periods with NA for missing values
clean_data <- raw_data %>%
  mutate(across(
    c(Deaths, `Death Rate`, `Age Adjusted Death Rate`),
    ~ na_if(.x, ".")
  ))

# Adjusting data types
clean_data <- clean_data %>%
  mutate(
    Year = as.integer(Year),
    Deaths = as.numeric(Deaths),
    Death_Rate = as.numeric(`Death Rate`),
    Age_Adjusted_Death_Rate = as.numeric(`Age Adjusted Death Rate`)
  )

write_path <- here("data", "analysis_data", "clean_nyc_data.csv")

write_csv(clean_data, write_path)
```

### 2024-03-15 scripts/00-simulated_dataset.r

@@ -1,0 +1,53 @@
+
+# Load libraries
+library(dplyr)
+library(tidyr)
+library(stringr)
+library(testthat)
+
+set.seed(853) # Set a seed for reproducibility
+
+
+# Number of rows to simulate
+n <- 500
+
+# Leading causes of death and their ICD-10 codes
+leading_causes <- c(
+  "Human immunodeficiency virus (HIV) disease B20-B24",
+  "Malignant neoplasms C00-C97",
+  "Diabetes mellitus E10-E14",
+  "Alzheimer’s disease G30",
+  "Diseases of heart I00-I09,I11,I13,I20-I51",
+  "Essential hypertension and hypertensive renal disease I10, I12",
+  "Cerebrovascular diseases I60-I69",
+  "Atherosclerosis I70",
+  "Other diseases of circulatory system I71-I78",
+  "Influenza and pneumonia J10-J18",
+  "Chronic lower respiratory diseases J40-J47",
+  "Chronic liver disease and cirrhosis K70,K73-K74",
+  "Nephritis, nephrotic syndrome and nephrosis N00-N07,N17-N19,N25-N27",
+  "Certain conditions originating in the perinatal period P00-P96",
+  "Congenital malformations, deformations & chromosomal abnormalities Q00-Q99",
+  "Sudden infant death syndrome R95",
+  "Symptoms, signs and abnormal clinical and laboratory findings, not elsewhere classified R00-R94,R96-R99",
+  "All other diseases (Residual)",
+  "Motor vehicle accidents V02-V04,V09.0,V09.2,V12-V14,V19.0-V19.2,V19.4-V19.6,V20-V79,V80.3-V80.5,V81.0-V81.1,V82.0-V82.1,V83-V86,V87.0-V87.8,V88.0-V88.8,V89.0,V89.2",
+  "All other and unspecified accidents and adverse effects V01,V05-V06,V09.1,V09.3-V09.9,V10-V11,V15-V18,V19.3,V19.8-V19.9,V80.0-V80.2,V80.6-V80.9,V81.2-V81.9,V82.2-V82.9,V87.9,V88.9,V89.1,V89.3,V89.9,V90-X59,Y40-Y86,Y88",
+  "Intentional self-harm (suicide) X60-X84,Y87.0",
+  "Assault (homicide) X85-Y09,Y87.1",
+  "All other external causes Y10-Y36,Y87.2,Y89"
+)
+
+
+# Generate the data
[…]

### 2024-03-15 scripts/02-cleaned_data.R

@@ -1,0 +1,37 @@
+
+# load libraries
+library(tidyverse)
+library(here)
+
+
+#Reading the raw data
+
+file_path <-
+  here("data",
+       "raw_data",
+       "New_York_City_Leading_Causes_of_Death_20240306.csv")
+raw_data <- read_csv(file_path)
+
+
+# Replacing periods with NA for missing values
+clean_data <- raw_data %>%
+  mutate(across(
+    c(Deaths, `Death Rate`, `Age Adjusted Death Rate`),
+    ~ na_if(.x, ".")
+  ))
+
+# Adjusting data types
+clean_data <- clean_data %>%
+  mutate(
+    Year = as.integer(Year),
+    Deaths = as.numeric(Deaths),
+    Death_Rate = as.numeric(`Death Rate`),
+    Age_Adjusted_Death_Rate = as.numeric(`Age Adjusted Death Rate`)
+  )
+
+
+write_path <- here("data", "analysis_data", "clean_nyc_data.csv")
+
+write_csv(clean_data, write_path)
+
+

### 2024-03-15 scripts/03-test_data.R

@@ -1,0 +1,52 @@
+# Load libraries 
+library(testthat)
+library(dplyr)
+library(tidyr)
+suppressPackageStartupMessages(library(stringr))
+
+
+# Script #1 to be tested - simulated data
+source("00-simulated_dataset.r", local = TRUE)
+
+# Ensure the simulated dataset has properties similar to the original dataset
+test_that("Data has correct types", {
+  expect_is(simulated_data$Year, "integer")
+  expect_is(simulated_data$Leading_Cause, "character")
+  expect_is(simulated_data$Sex, "character")
+  expect_is(simulated_data$Race_Ethnicity, "character")
+  expect_is(simulated_data$Deaths, "integer")
+  expect_is(simulated_data$Death_Rate, "numeric")
+  expect_is(simulated_data$Age_Adjusted_Death_Rate, "numeric")
+})
+
+# Ensure 'Year' column contains only the years in the specified range
+test_that("Year column contains appropriate years", {
+  expect_true(all(simulated_data$Year >= 2007 & simulated_data$Year <= 2014))
+})
+
+# Ensure no duplicated rows exists
+test_that("There are no duplicate rows", {
+  expect_equal(nrow(simulated_data), nrow(distinct(simulated_data)))
+})
+
+
+# Script #2 to be tested - causes of death
+source("04-causes_of_death_results.R", local = TRUE)
+
+# Data set has been loaded correctly
+test_that("Dataset has been loaded", {
+  expect_true(exists("nyc_data"))
+  expect_is(nyc_data, "data.frame")
+})
+
+# 'death_count_by_cause' has the correct columns
+test_that("'death_count_by_cause' has correct columns", {
+  expected_cols <- c("Leading.Cause", "Total_Deaths")
+  expect_equal(colnames(death_count_by_cause), expected_cols)
+})
+
+# 'death_count_by_cause' table has correct number of rows
+test_that("'death_count_by_cause' has correct number of rows", {
+  expect_true(nrow(death_count_by_cause) > 0)
+})
+

### 2024-03-15 scripts/04-causes_of_death_results.R

@@ -1,0 +1,69 @@
+
+# libraries
+library(dplyr)
+library(knitr)
+library(ggplot2)
+library(forcats)
+library(testthat)
+
+# read in the dataset
+
+file_path <- "../data/analysis_data/clean_nyc_data.csv"
+nyc_data <- read.csv(file_path)
+
+
+# create table showing count of deaths by cause (overall)
+death_count_by_cause <- nyc_data %>%
+  group_by(Leading.Cause) %>%
+  summarise(Total_Deaths = sum(as.numeric(Deaths), na.rm = TRUE)) %>%
+  arrange(desc(Total_Deaths))
+knitr::kable(death_count_by_cause, caption = "Count of Deaths by Leading Cause")
+
+
+# create table showing top 5 leading causes of death by race
+top_causes_by_race <- nyc_data %>%
+  group_by(Race.Ethnicity, Leading.Cause) %>%
+  summarise(Total_Deaths = sum(as.numeric(Deaths), na.rm = TRUE)) %>%
+  arrange(Race.Ethnicity, desc(Total_Deaths)) %>%
+  mutate(rank = row_number()) %>%
+  filter(rank <= 5) %>%
+  ungroup()
+
+table_top_causes_by_race <- top_causes_by_race %>%
+  select(-rank) %>%
+  arrange(Race.Ethnicity, desc(Total_Deaths))
+
+knitr::kable(table_top_causes_by_race, caption = "Top 5 Leading Causes of Death by Race")
+
+# create bar graph to display leading causes of death by race
+gg <- ggplot(top_causes_by_race, aes(x = Leading.Cause, y = Total_Deaths)) +
+  geom_col(aes(fill = Leading.Cause)) + 
+  facet_wrap(~Race.Ethnicity, scales = "free_y") + 
+  scale_fill_brewer(palette = "Paired") + 
+  theme_minimal() + 
+  theme(axis.text.x = element_text(angle = 40, hjust = 1), 
+        strip.text.x = element_text(size = 6), 
+        legend.position = "none") + 
+  labs(x = "Top 5 Causes of Death", y = "Total Deaths") 
+
+print(gg)
+ggsave("top_causes_by_race.png", plot = gg, width = 12, height = 8)
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+

### 2024-03-15 paper/paper.qmd
rendered paper to latest versio

@@
 # Model
+
+In our research, we investigated the causes of mortality within New York City’s diverse racial and ethnic populations for the year 2014.
+We hypothesize that the total number of deaths by leading cause follows a negative binomial regression.
+Formally, the model is expressed as:
+$y_i | n_i \sim \text{NegBin}(\mu_i, k)$
+
+In the expression above, $y_i$, signifies the count of deaths for cause i,$\mu_i$ is the expected count, and $k$ is the dispersion parameter, which captures over-dispersion. 
+Our model also utilizes a log-link function, modeling the expected log-count as:
+$\log(\mu_i) = \beta_0 + \beta_1 \times \text{Cause}_i$
+
+Where $\beta_0$ is the intercept, and $\beta_1$ represents the vector of coefficients for the leading causes of death. 
+The leading causes are treated as categorical predictors, which results in the need to use dummy coding with one reference category.
+
+For our Bayesian regression framework, we employ the stan_glm() function from the rstanarm package, which assumes default normal priors for the regression coefficients as follows:
+$\beta_0$ ~ Normal(0,2.5)
+$\beta_1$ ~ Normal(0,2.5)
 

### 2024-03-16 scripts/03-test_data.R
added preamble

@@ -1,3 +1,10 @@
+#### Preamble ####
+# Purpose: Tests for simulated dataset and causes of death files 
+# Author: Maria Mangru, Rayan Awad Alim, MD Mubtasim-Faud
+# Date: 15th March, 2024
+# License: MIT
+
+
 # Load libraries 
 library(testthat)
 library(dplyr)

### 2024-03-16 scripts/04-causes_of_death_results.R

@@ -1,3 +1,10 @@
+#### Preamble ####
+# Purpose: Generates graphs and tables showing the causes of death
+# Author: Maria Mangru, Rayan Awad Alim, MD Mubtasim-Faud
+# Date: 15th March, 2024
+# License: MIT
+
+
 
 # libraries
 library(dplyr)
@@ -8,7 +15,7 @@ library(testthat)
 
 # read in the dataset
 
-file_path <- "../data/analysis_data/clean_nyc_data.csv"
+file_path <- "./data/analysis_data/clean_nyc_data.csv"
 nyc_data <- read.csv(file_path)
 
 
@@ -49,6 +56,9 @@ gg <- ggplot(top_causes_by_race, aes(x = Leading.Cause, y = Total_Deaths)) +
 print(gg)
 ggsave("top_causes_by_race.png", plot = gg, width = 12, height = 8)
 
+# Save datatsets for use in models 
+write.csv(death_count_by_cause, "data/analysis_data/death_count_by_cause.csv", row.names = FALSE)
+write.csv(table_top_causes_by_race, "data/analysis_data/top_causes_by_race.csv", row.names = FALSE)
 
 
 

### 2024-03-16 paper/paper.qmd
update paper visualisations

@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE}
 #| include: false
@@
 
+# Load libraries
 library(tidyverse)
-library(palmerpenguins)
+library(dplyr)
+library(knitr)
+library(ggplot2)
+library(forcats)
+library(testthat)
+library(kableExtra)
+
 ```
@@
 
-# Data {#sec-data}
+# Data
 
-DATA This analysis utilizes data from the New York City Department of Health and Mental Hygiene, specifically provided by the Bureau of Vital Statistics {@NYCData}.
+This analysis utilizes data from the New York City Department of Health and Mental Hygiene, specifically provided by the Bureau of Vital Statistics {@NYCData}.
 The dataset encompasses records of deaths in New York City since 2007, detailing the leading causes of death categorized by sex and ethnicity.
@@
 
-Dataset Context and Broader Implications The investigation into the leading causes of death within New York City holds significant public health importance.
+## Dataset Context and Broader Implications
+
+The investigation into the leading causes of death within New York City holds significant public health importance.
 It allows for a nuanced understanding of mortality trends, guiding health policy, and intervention strategies tailored to specific demographics and causes.
@@
 
-Variables and Data Examination The dataset comprises several key variables, including: Year: The year of the recorded death.
-Leading Cause: The primary cause of death as categorized by ICD-10 codes.
-Sex: The sex of the deceased (Male, Female, or Gender Non-Conforming).
-Race Ethnicity: The self-reported ethnicity of the deceased, including categories such as Hispanic, White Non-Hispanic, Black Non-Hispanic, Asian and Pacific Islander, and Others.
-Deaths: The total number of deaths attributed to the leading cause.
-Death Rate and Age-Adjusted Death Rate: Rates per 100,000 population, providing standardized comparisons across different population sizes and age structures.
+## Variables and Data Examination 
+The dataset comprises several key variables, including: 
+- Year: The year of the recorded death.
 
-Alternative Datasets and Justification While other mortality databases exist, such as the CDC's National Center for Health Statistics, the chosen dataset offers a detailed and localized perspective specific to New York City.
+- Leading Cause: The primary cause of death as categorized by ICD-10 codes.
+
+- Sex: The sex of the deceased (Male, Female, or Gender Non-Conforming).
+
+- Race Ethnicity: The self-reported ethnicity of the deceased, including categories such as Hispanic, White, Non-Hispanic, Black Non-Hispanic, Asian and Pacific Islander, and others.
+
+- Deaths: The total number of deaths attributed to the leading cause.
+
+- Death Rate and Age-Adjusted Death Rate: Rates per 100,000 population, providing standardized comparisons across different population sizes and age structures.
+
+## Alternative Datasets and Justification 
+While other mortality databases exist, such as the CDC's National Center for Health Statistics, the chosen dataset offers a detailed and localized perspective specific to New York City.
 This specificity provides a more precise tool for understanding and addressing urban health dynamics, making it more relevant for city-specific policy-making and health intervention strategies.
 
-Data Cleaning and Variable Construction Preliminary data cleaning focused on addressing missing values, ensuring consistency in categorical variables, and suppressing unreliable rates as mentioned.
+## Data Cleaning and Variable Construction
+
+Preliminary data cleaning focused on addressing missing values, ensuring consistency in categorical variables, and suppressing unreliable rates as mentioned.
 New variables, such as aggregated categories for cause of death or demographic groupings, were constructed to facilitate analysis.
@@
 
-Summary Statistics and Relationships Initial exploratory data analysis revealed key trends and disparities in mortality rates across different populations within New York City.
+## Summary Statistics and Relationships 
+Initial exploratory data analysis revealed key trends and disparities in mortality rates across different populations within New York City.
 We provide summary statistics, including mean death rates and distributions of deaths by cause ethnicity, to provide insights into the health landscape of the city.
@@
 
-Measurement and Methodological Notes The measurement of mortality and its causes relies on the accurate classification of death certificates, adhering to ICD-10 standards.
+## Measurement and Methodology Notes
+
+The measurement of mortality and its causes relies on the accurate classification of death certificates, adhering to ICD-10 standards.
 This classification ensures comparability with other datasets and robustness in identifying health trends.
@@
 Formally, the model is expressed as:
+
 $y_i | n_i \sim \text{NegBin}(\mu_i, k)$
@@
 Our model also utilizes a log-link function, modeling the expected log-count as:
+
 $\log(\mu_i) = \beta_0 + \beta_1 \times \text{Cause}_i$
@@
 ### Model justification
+The negative binomial regression was used due to its ability to account for the over-dispersion present in the data.
+This is a common occurrence with mortality counts due to the heterogeneity of death causes.
+Through the use of the  parameter, we can model the variance separately from the mean which provides a more flexible and accurate representation of the data.
+This model also accommodates the count nature of the dependent variable.
+Furthermore, the model’s priors are informed by the normative distributions typically used in Bayesian regression analysis.
+This enables a degree of regularization that reduces the potential for overfitting.
 
 # Results
+
+```{r, echo=FALSE, message=FALSE}
+file_path <- "../data/analysis_data/clean_nyc_data.csv"
+nyc_data <- read.csv(file_path)
+
+# create table showing count of deaths by cause (overall)
+death_count_by_cause <- nyc_data %>%
+  group_by(Leading.Cause) %>%
+  summarise(Total_Deaths = sum(as.numeric(Deaths), na.rm = TRUE)) %>%
+  arrange(desc(Total_Deaths)) %>%
+  slice_head(n = 10) %>%
+  mutate(Row = row_number()) %>%
+  select(Row, everything())
+
+colnames(death_count_by_cause) <-
+  c("No.", "Cause of Death", "Total Death Count")
+
+knitr::kable(death_count_by_cause, caption = "Count of Deaths by Leading Cause") %>%
+  kable_styling(bootstrap_options = c("striped", "hover"),
+                full_width = F) %>%
+  column_spec(1, bold = TRUE) %>%
+  column_spec(2, bold = TRUE, color = "blue") %>%
+  column_spec(3, width = "3cm")
+
+```
+\newpage
+
+```{r, echo=FALSE, message=FALSE, fig.cap="Top 5 Causes of Death by Race. This graph displays the leading causes of death, broken down by race/ethnicity", fig.width=10, fig.height=8, echo=FALSE}
+# create table showing top 5 leading causes of death by race
+top_causes_by_race <- nyc_data %>%
+  group_by(Race.Ethnicity, Leading.Cause) %>%
+  summarise(Total_Deaths = sum(as.numeric(Deaths), na.rm = TRUE)) %>%
+  arrange(Race.Ethnicity, desc(Total_Deaths)) %>%
+  mutate(rank = row_number()) %>%
+  filter(rank <= 5) %>%
+  ungroup()
+
+# create bar graph to display leading causes of death by race
+gg <- ggplot(top_causes_by_race, aes(x = Leading.Cause, y = Total_Deaths)) +
+  geom_col(aes(fill = Leading.Cause)) + 
+  facet_wrap(~Race.Ethnicity, scales = "free_y") + 
+  scale_fill_brewer(palette = "Paired") + 
+  theme_minimal() + 
+  theme(axis.text.x = element_text(angle = 40, hjust = 1), 
+        strip.text.x = element_text(size = 6), 
+        legend.position = "none") + 
+  labs(x = "Top 5 Causes of Death", y = "Total Deaths")
+
+print(gg)
+```
 
@@
 
-# Model details {#sec-model-details}
+# Model details
 

### 2024-03-16 scripts/model_asianpacific.R
added models

@@ -1,0 +1,49 @@
+# load libraries
+library(dplyr)
+library(readr)
+library(ggplot2)
+library(forcats)
+library(rstanarm)
+library(knitr)
+library(kableExtra)
+
+# load the dataset 
+file_path <- "../data/analysis_data/clean_nyc_data.csv"
+nyc_data <- read.csv(file_path)
+
+
+# Convert Deaths to numeric if it's not already
+nyc_data$Deaths <- as.numeric(as.character(nyc_data$Deaths))
+
+# Filter for the year 2014 and the specific race
+race_name <- "Asian and Pacific Islander" 
+
+nyc_data_2014_race <- nyc_data %>%
+  filter(Year == 2014, Race.Ethnicity == race_name)
+
+# Identify the top 5 causes of death for the selected race
+top_causes_by_race <- nyc_data_2014_race %>%
+  group_by(Leading.Cause) %>%
+  summarise(Total_Deaths = sum(as.numeric(Deaths), na.rm = TRUE), .groups = 'drop') %>%
+  arrange(desc(Total_Deaths)) %>%
+  slice_max(Total_Deaths, n = 5)
+
+# Ensure data for regression includes only the top causes
+nyc_data_for_regression <- nyc_data_2014_race %>%
+  inner_join(top_causes_by_race, by = "Leading.Cause")
+
+
+# negative binomial regression
+neg_binom_model <- stan_glm(
+  Total_Deaths ~ Leading.Cause,
+  data = nyc_data_for_regression,
+  family = neg_binomial_2(link = "log"),
+  seed = 853
+)
+
+# Output the model summary
+model_summary <- summary(neg_binom_model)
+
+# Display the model summary table
+kable(model_summary, caption = paste("Negative Binomial Regression Summary for", race_name)) %>%
+  kable_styling(bootstrap_options = c("striped", "hover"))

### 2024-03-16 scripts/model_black.R

@@ -1,0 +1,49 @@
+# load libraries
+library(dplyr)
+library(readr)
+library(ggplot2)
+library(forcats)
+library(rstanarm)
+library(knitr)
+library(kableExtra)
+
+# load the dataset 
+file_path <- "../data/analysis_data/clean_nyc_data.csv"
+nyc_data <- read.csv(file_path)
+
+
+# Convert Deaths to numeric if it's not already
+nyc_data$Deaths <- as.numeric(as.character(nyc_data$Deaths))
+
+# Filter for the year 2014 and the specific race
+race_name <- "Black Non-Hispanic" 
+
+nyc_data_2014_race <- nyc_data %>%
+  filter(Year == 2014, Race.Ethnicity == race_name)
+
+# Identify the top 5 causes of death for the selected race
+top_causes_by_race <- nyc_data_2014_race %>%
+  group_by(Leading.Cause) %>%
+  summarise(Total_Deaths = sum(as.numeric(Deaths), na.rm = TRUE), .groups = 'drop') %>%
+  arrange(desc(Total_Deaths)) %>%
+  slice_max(Total_Deaths, n = 5)
+
+# Ensure data for regression includes only the top causes
+nyc_data_for_regression <- nyc_data_2014_race %>%
+  inner_join(top_causes_by_race, by = "Leading.Cause")
+
+
+# negative binomial regression
+neg_binom_model <- stan_glm(
+  Total_Deaths ~ Leading.Cause,
+  data = nyc_data_for_regression,
+  family = neg_binomial_2(link = "log"),
+  seed = 853
+)
+
+# Output the model summary
+model_summary <- summary(neg_binom_model)
+
+# Display the model summary table
+kable(model_summary, caption = paste("Negative Binomial Regression Summary for", race_name)) %>%
+  kable_styling(bootstrap_options = c("striped", "hover"))

### 2024-03-16 scripts/model_hispanic.R

@@ -1,0 +1,49 @@
+# load libraries
+library(dplyr)
+library(readr)
+library(ggplot2)
+library(forcats)
+library(rstanarm)
+library(knitr)
+library(kableExtra)
+
+# load the dataset 
+file_path <- "../data/analysis_data/clean_nyc_data.csv"
+nyc_data <- read.csv(file_path)
+
+
+# Convert Deaths to numeric if it's not already
+nyc_data$Deaths <- as.numeric(as.character(nyc_data$Deaths))
+
+# Filter for the year 2014 and the specific race
+race_name <- "Hispanic" 
+
+nyc_data_2014_race <- nyc_data %>%
+  filter(Year == 2014, Race.Ethnicity == race_name)
+
+# Identify the top 5 causes of death for the selected race
+top_causes_by_race <- nyc_data_2014_race %>%
+  group_by(Leading.Cause) %>%
+  summarise(Total_Deaths = sum(as.numeric(Deaths), na.rm = TRUE), .groups = 'drop') %>%
+  arrange(desc(Total_Deaths)) %>%
+  slice_max(Total_Deaths, n = 5)
+
+# Ensure data for regression includes only the top causes
+nyc_data_for_regression <- nyc_data_2014_race %>%
+  inner_join(top_causes_by_race, by = "Leading.Cause")
+
+
+# negative binomial regression
+neg_binom_model <- stan_glm(
+  Total_Deaths ~ Leading.Cause,
+  data = nyc_data_for_regression,
+  family = neg_binomial_2(link = "log"),
+  seed = 853
+)
+
+# Output the model summary
+model_summary <- summary(neg_binom_model)
+
+# Display the model summary table
+kable(model_summary, caption = paste("Negative Binomial Regression Summary for", race_name)) %>%
+  kable_styling(bootstrap_options = c("striped", "hover"))

### 2024-03-16 scripts/model_notstated.R

@@ -1,0 +1,49 @@
+# load libraries
+library(dplyr)
+library(readr)
+library(ggplot2)
+library(forcats)
+library(rstanarm)
+library(knitr)
+library(kableExtra)
+
+# load the dataset 
+file_path <- "../data/analysis_data/clean_nyc_data.csv"
+nyc_data <- read.csv(file_path)
+
+
+# Convert Deaths to numeric if it's not already
+nyc_data$Deaths <- as.numeric(as.character(nyc_data$Deaths))
+
+# Filter for the year 2014 and the specific race
+race_name <- "Not Stated/Unknown" 
+
+nyc_data_2014_race <- nyc_data %>%
+  filter(Year == 2014, Race.Ethnicity == race_name)
+
+# Identify the top 5 causes of death for the selected race
+top_causes_by_race <- nyc_data_2014_race %>%
+  group_by(Leading.Cause) %>%
+  summarise(Total_Deaths = sum(as.numeric(Deaths), na.rm = TRUE), .groups = 'drop') %>%
+  arrange(desc(Total_Deaths)) %>%
+  slice_max(Total_Deaths, n = 5)
+
+# Ensure data for regression includes only the top causes
+nyc_data_for_regression <- nyc_data_2014_race %>%
+  inner_join(top_causes_by_race, by = "Leading.Cause")
+
+
+# negative binomial regression
+neg_binom_model <- stan_glm(
+  Total_Deaths ~ Leading.Cause,
+  data = nyc_data_for_regression,
+  family = neg_binomial_2(link = "log"),
+  seed = 853
+)
+
+# Output the model summary
+model_summary <- summary(neg_binom_model)
+
+# Display the model summary table
+kable(model_summary, caption = paste("Negative Binomial Regression Summary for", race_name)) %>%
+  kable_styling(bootstrap_options = c("striped", "hover"))

### 2024-03-16 scripts/model_other.R

@@ -1,0 +1,48 @@
+# load libraries
+library(dplyr)
+library(readr)
+library(ggplot2)
+library(forcats)
+library(rstanarm)
+library(knitr)
+library(kableExtra)
+
+# load the dataset 
+file_path <- "../data/analysis_data/clean_nyc_data.csv"
+nyc_data <- read.csv(file_path)
+
+# Convert Deaths to numeric if it's not already
+nyc_data$Deaths <- as.numeric(as.character(nyc_data$Deaths))
+
+# Filter for the year 2014 and the specific race
+race_name <- "Other Race/Ethnicity" 
+
+nyc_data_2014_race <- nyc_data %>%
+  filter(Year == 2014, Race.Ethnicity == race_name)
+
+# Identify the top 5 causes of death for the selected race
+top_causes_by_race <- nyc_data_2014_race %>%
+  group_by(Leading.Cause) %>%
+  summarise(Total_Deaths = sum(as.numeric(Deaths), na.rm = TRUE), .groups = 'drop') %>%
+  arrange(desc(Total_Deaths)) %>%
+  slice_max(Total_Deaths, n = 5)
+
+# Ensure data for regression includes only the top causes
+nyc_data_for_regression <- nyc_data_2014_race %>%
+  inner_join(top_causes_by_race, by = "Leading.Cause")
+
+
+# negative binomial regression
+neg_binom_model <- stan_glm(
+  Total_Deaths ~ Leading.Cause,
+  data = nyc_data_for_regression,
+  family = neg_binomial_2(link = "log"),
+  seed = 853
+)
+
+# Output the model summary
+model_summary <- summary(neg_binom_model)
+
+# Display the model summary table
+kable(model_summary, caption = paste("Negative Binomial Regression Summary for", race_name)) %>%
+  kable_styling(bootstrap_options = c("striped", "hover"))

### 2024-03-16 scripts/model_overall_causes.R

@@ -1,0 +1,41 @@
+# load libraries
+library(dplyr)
+library(readr)
+library(ggplot2)
+library(forcats)
+library(rstanarm)
+library(knitr)
+library(kableExtra)
+
+# load the dataset 
+file_path <- "../data/analysis_data/clean_nyc_data.csv"
+nyc_data <- read.csv(file_path)
+
+
+# Filter for the year 2014
+nyc_data$Deaths <- as.numeric(as.character(nyc_data$Deaths))
+nyc_data_2014 <- nyc_data %>%
+  filter(Year == 2014)
+
+# Identify top 5 causes of death overall
+top_causes_overall <- nyc_data_2014 %>%
+  group_by(Leading.Cause) %>%
+  summarise(Total_Deaths = sum(Deaths, na.rm = TRUE)) %>%
+  ungroup() %>%
+  arrange(desc(Total_Deaths)) %>%
+  slice_max(order_by = Total_Deaths, n = 5)
+
+# Filter data for top 5 causes
+nyc_data_2014_top_causes_with_totals <- nyc_data_2014 %>%
+  inner_join(top_causes_overall, by = "Leading.Cause")
+
+# Negative binomial regression for overall top 5 causes
+neg_binom_model_overall <- stan_glm(
+  Total_Deaths ~ Leading.Cause,
+  data = nyc_data_2014_top_causes_with_totals,
+  family = neg_binomial_2(link = "log"),
+  seed = 853
+)
+
+# Summary model
+summary(neg_binom_model_overall) 

### 2024-03-16 scripts/model_white.R

@@ -1,0 +1,49 @@
+
+# load libraries
+library(dplyr)
+library(readr)
+library(ggplot2)
+library(forcats)
+library(rstanarm)
+library(knitr)
+library(kableExtra)
+
+# load the dataset 
+file_path <- "../data/analysis_data/clean_nyc_data.csv"
+nyc_data <- read.csv(file_path)
+
+
+# Convert Deaths to numeric if it's not already
+nyc_data$Deaths <- as.numeric(as.character(nyc_data$Deaths))
+
+# Filter for the year 2014 and the specific race
+race_name <- "White Non-Hispanic" 
+
+nyc_data_2014_race <- nyc_data %>%
+  filter(Year == 2014, Race.Ethnicity == race_name)
+
+# Identify the top 5 causes of death for the selected race
+top_causes_by_race <- nyc_data_2014_race %>%
+  group_by(Leading.Cause) %>%
+  summarise(Total_Deaths = sum(as.numeric(Deaths), na.rm = TRUE), .groups = 'drop') %>%
+  arrange(desc(Total_Deaths)) %>%
+  slice_max(Total_Deaths, n = 5)
+
+# Ensure data for regression includes only the top causes
+nyc_data_for_regression <- nyc_data_2014_race %>%
+  inner_join(top_causes_by_race, by = "Leading.Cause")
+
+# negative binomial regression
+neg_binom_model <- stan_glm(
+  Total_Deaths ~ Leading.Cause,
+  data = nyc_data_for_regression,
+  family = neg_binomial_2(link = "log"),
+  seed = 853
+)
+
+# Output the model summary
+model_summary <- summary(neg_binom_model)
+
+# Display the model summary table
+kable(model_summary, caption = paste("Negative Binomial Regression Summary for", race_name)) %>%
+  kable_styling(bootstrap_options = c("striped", "hover"))

### 2024-03-16 scripts/model_asianpacific.R
added preambles

@@ -1,3 +1,9 @@
+#### Preamble ####
+# Purpose: Regression model for causes of death for asian pacific  
+# Author: Maria Mangru, Rayan Awad Alim, MD Mubtasim-Faud
+# Date: 15th March, 2024
+# License: MIT
+
 # load libraries
 library(dplyr)
 library(readr)

### 2024-03-16 scripts/model_black.R

@@ -1,3 +1,9 @@
+#### Preamble ####
+# Purpose: Regression model for causes of death for black  
+# Author: Maria Mangru, Rayan Awad Alim, MD Mubtasim-Faud
+# Date: 15th March, 2024
+# License: MIT
+
 # load libraries
 library(dplyr)
 library(readr)

### 2024-03-16 scripts/model_hispanic.R

@@ -1,3 +1,9 @@
+#### Preamble ####
+# Purpose: Regression model for causes of death for hispanic  
+# Author: Maria Mangru, Rayan Awad Alim, MD Mubtasim-Faud
+# Date: 15th March, 2024
+# License: MIT
+
 # load libraries
 library(dplyr)
 library(readr)

### 2024-03-16 scripts/model_notstated.R

@@ -1,3 +1,9 @@
+#### Preamble ####
+# Purpose: Regression model for causes of death for not stated  
+# Author: Maria Mangru, Rayan Awad Alim, MD Mubtasim-Faud
+# Date: 15th March, 2024
+# License: MIT
+
 # load libraries
 library(dplyr)
 library(readr)

### 2024-03-16 scripts/model_other.R

@@ -3,6 +3,12 @@ library(dplyr)
 library(readr)
 library(ggplot2)
 library(forcats)
+#### Preamble ####
+# Purpose: Regression model for causes of death for other 
+# Author: Maria Mangru, Rayan Awad Alim, MD Mubtasim-Faud
+# Date: 15th March, 2024
+# License: MIT
+
 library(rstanarm)
 library(knitr)
 library(kableExtra)

### 2024-03-16 scripts/model_overall_causes.R

@@ -1,3 +1,10 @@
+#### Preamble ####
+# Purpose: Regression model for causes of death for overall population  
+# Author: Maria Mangru, Rayan Awad Alim, MD Mubtasim-Faud
+# Date: 15th March, 2024
+# License: MIT
+
+
 # load libraries
 library(dplyr)
 library(readr)

### 2024-03-16 scripts/model_white.R

@@ -1,3 +1,9 @@
+#### Preamble ####
+# Purpose: Regression model for causes of death for white non hispanic  
+# Author: Maria Mangru, Rayan Awad Alim, MD Mubtasim-Faud
+# Date: 15th March, 2024
+# License: MIT
+
 
 # load libraries
 library(dplyr)

### 2024-03-16 models/model_asianpacific.Rmd
Model validation, comparision

---
title: "model_asianpacific"
format: html
---

```{r}
# load libraries
library(dplyr)
library(readr)
library(ggplot2)
library(forcats)
library(rstanarm)
library(knitr)
library(kableExtra)
```

```{r}
# load the dataset 
file_path <- "../data/analysis_data/clean_nyc_data.csv"
nyc_data <- read.csv(file_path)
```

```{r}
# Convert Deaths to numeric if it's not already
nyc_data$Deaths <- as.numeric(as.character(nyc_data$Deaths))

# Filter for the year 2014 and the specific race
race_name <- "Asian and Pacific Islander"

nyc_data_2014_race <- nyc_data %>%
  filter(Year == 2014, Race.Ethnicity == race_name)

# Identify the top 5 causes of death for the selected race
top_causes_by_race <- nyc_data_2014_race %>%
  group_by(Leading.Cause) %>%
  summarise(Total_Deaths = sum(as.numeric(Deaths), na.rm = TRUE),
            .groups = 'drop') %>%
  arrange(desc(Total_Deaths)) %>%
  slice_max(Total_Deaths, n = 5)

# Ensure data for regression includes only the top causes
nyc_data_for_regression <- nyc_data_2014_race %>%
  inner_join(top_causes_by_race, by = "Leading.Cause")
```

```{r}
# negative binomial regression
neg_binom_model <- stan_glm(
  Total_Deaths ~ Leading.Cause,
  data = nyc_data_for_regression,
  family = neg_binomial_2(link = "log"),
  seed = 853
)

# Output the model summary
model_summary <- summary(neg_binom_model)

# Display the model summary table
kable(model_summary,
      caption = paste("Negative Binomial Regression Summary for", race_name)) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```
```{r}
# Fit the Poisson model
poisson_model <- stan_glm(
  Total_Deaths ~ Leading.Cause,
  data = nyc_data_for_regression,
  family = poisson(link = "log"),
  seed = 853
)

# Output the model summary
model_summary_poi <- summary(poisson_model)

# Display the model summary table
kable(model_summary_poi, caption = paste("Poission Summary for", race_name)) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r}
# Leave-One-Out Cross-Validation (LOO) for model comparison
loo_neg_binom_asian <- loo(neg_binom_model)
loo_poisson_asian <- loo(poisson_model)
loo_comparison_asian <- loo_compare(loo_neg_binom_asian, loo_poisson_asian)
print(loo_comparison_asian)
```

```{r}
# Posterior Predictive Checks
# Posterior predictive check for the Negative Binomial model
pp_check_neg_binom_asian <-
  pp_check(neg_binom_model)  + ggtitle("Negative Binomial Model")
ggsave(
  "../data/analysis_data/Model_Output/pp_check_neg_binom_asian.png",
  plot = pp_check_neg_binom_asian,
  width = 10,
  height = 8
)
pp_check_neg_binom_asian

# Posterior predictive check for the Poisson model
pp_check_poisson_asian <-
  pp_check(poisson_model)  + ggtitle("Poisson Model")
ggsave(
  "../data/analysis_data/Model_Output/pp_check_poisson_asian.png",
  plot = pp_check_poisson_asian,
  width = 10,
  height = 8
)
pp_check_poisson_asian
```

### 2024-03-16 models/model_black.Rmd
Model validation, comparision

--- 2024-03-09 scripts/model_asianpacific.Rmd
@@
 ---
-title: "model_asianpacific"
+title: "model_black"
 format: html
@@
 # Filter for the year 2014 and the specific race
-race_name <- "Asian and Pacific Islander"
+race_name <- "Black Non-Hispanic"
 
@@
 ```
+
+```{r}
+# Leave-One-Out Cross-Validation (LOO) for model comparison
+loo_neg_binom_black <- loo(neg_binom_model)
+loo_poisson_black <- loo(poisson_model)
+loo_comparison_black <- loo_compare(loo_neg_binom_black, loo_poisson_black)
+print(loo_comparison_black)
+```
+
+```{r}
+# Posterior Predictive Checks
+# Posterior predictive check for the Negative Binomial model
+pp_check_neg_binom_black <-
+  pp_check(neg_binom_model)  + ggtitle("Negative Binomial Model")
+ggsave(
+  "../data/analysis_data/Model_Output/pp_check_neg_binom_black.png",
+  plot = pp_check_neg_binom_black,
+  width = 10,
+  height = 8
+)
+pp_check_neg_binom_black
+
+# Posterior predictive check for the Poisson model
+pp_check_poisson_black <-
+  pp_check(poisson_model)  + ggtitle("Poisson Model")
+ggsave(
+  "../data/analysis_data/Model_Output/pp_check_poisson_black.png",
+  plot = pp_check_poisson_black,
+  width = 10,
+  height = 8
+)
+pp_check_poisson_black
+```

### 2024-03-16 models/model_hispanic.Rmd
moving files
moved the model files to the appropriate folder.
Model validation, comparision

--- 2024-03-09 scripts/model_asianpacific.Rmd
@@
 ---
-title: "model_asianpacific"
+title: "model_hispanic"
 format: html
@@
 # Filter for the year 2014 and the specific race
-race_name <- "Asian and Pacific Islander"
+race_name <- "Hispanic"
 
@@
 ```
+
+```{r}
+# Leave-One-Out Cross-Validation (LOO) for model comparison
+loo_neg_binom_hispanic <- loo(neg_binom_model)
+loo_poisson_hispanic <- loo(poisson_model)
+loo_comparison_hispanic <- loo_compare(loo_neg_binom_hispanic, loo_poisson_hispanic)
+print(loo_comparison_hispanic)
+```
+
+```{r}
+# Posterior Predictive Checks
+# Posterior predictive check for the Negative Binomial model
+pp_check_neg_binom_hispanic <-
+  pp_check(neg_binom_model)  + ggtitle("Negative Binomial Model")
+ggsave(
+  "../data/analysis_data/Model_Output/pp_check_neg_binom_hispanic.png",
+  plot = pp_check_neg_binom_hispanic,
+  width = 10,
+  height = 8
+)
+pp_check_neg_binom_hispanic
+
+# Posterior predictive check for the Poisson model
+pp_check_poisson_hispanic <-
+  pp_check(poisson_model)  + ggtitle("Poisson Model")
+ggsave(
+  "../data/analysis_data/Model_Output/pp_check_poisson_hispanic.png",
+  plot = pp_check_poisson_hispanic,
+  width = 10,
+  height = 8
+)
+pp_check_poisson_hispanic
+```

### 2024-03-16 models/model_notstated.Rmd
Model validation, comparision

--- 2024-03-09 scripts/model_asianpacific.Rmd
@@
 ---
-title: "model_asianpacific"
+title: "model_notstated"
 format: html
@@
 # Filter for the year 2014 and the specific race
-race_name <- "Asian and Pacific Islander"
+race_name <- "Not Stated/Unknown"
 
@@
 ```
+
+```{r}
+# Leave-One-Out Cross-Validation (LOO) for model comparison
+loo_neg_binom_notstated <- loo(neg_binom_model)
+loo_poisson_notstated <- loo(poisson_model)
+loo_comparison_notstated <- loo_compare(loo_neg_binom_notstated, loo_poisson_notstated)
+print(loo_comparison_notstated)
+```
+
+```{r}
+# Posterior Predictive Checks
+# Posterior predictive check for the Negative Binomial model
+pp_check_neg_binom_notstated <-
+  pp_check(neg_binom_model)  + ggtitle("Negative Binomial Model")
+ggsave(
+  "../data/analysis_data/Model_Output/pp_check_neg_binom_notstated.png",
+  plot = pp_check_neg_binom_notstated,
+  width = 10,
+  height = 8
+)
+pp_check_neg_binom_notstated
+
+# Posterior predictive check for the Poisson model
+pp_check_poisson_notstated <-
+  pp_check(poisson_model)  + ggtitle("Poisson Model")
+ggsave(
+  "../data/analysis_data/Model_Output/pp_check_poisson_notstated.png",
+  plot = pp_check_poisson_notstated,
+  width = 10,
+  height = 8
+)
+pp_check_poisson_notstated
+```

### 2024-03-16 models/model_white.Rmd
moving files
moved the model files to the appropriate folder.
Model validation, comparision

--- 2024-03-09 scripts/model_asianpacific.Rmd
@@
 ---
-title: "model_asianpacific"
+title: "model_white"
 format: html
@@
 # Filter for the year 2014 and the specific race
-race_name <- "Asian and Pacific Islander"
+race_name <- "White Non-Hispanic"
 
@@
 ```
+
+```{r}
+# Leave-One-Out Cross-Validation (LOO) for model comparison
+loo_neg_binom_white <- loo(neg_binom_model)
+loo_poisson_white <- loo(poisson_model)
+loo_comparison_white <- loo_compare(loo_neg_binom_white, loo_poisson_white)
+print(loo_comparison_white)
+```
+
+```{r}
+# Posterior Predictive Checks
+# Posterior predictive check for the Negative Binomial model
+pp_check_neg_binom_white <-
+  pp_check(neg_binom_model)  + ggtitle("Negative Binomial Model")
+ggsave(
+  "../data/analysis_data/Model_Output/pp_check_neg_binom_white.png",
+  plot = pp_check_neg_binom_white,
+  width = 10,
+  height = 8
+)
+pp_check_neg_binom_white
+
+# Posterior predictive check for the Poisson model
+pp_check_poisson_white <-
+  pp_check(poisson_model)  + ggtitle("Poisson Model")
+ggsave(
+  "../data/analysis_data/Model_Output/pp_check_poisson_white.png",
+  plot = pp_check_poisson_white,
+  width = 10,
+  height = 8
+)
+pp_check_poisson_white
+```

### 2024-03-16 paper/paper.qmd
Update paper, clean up redundant files

@@
 ```
+
+## Model Results
+
+For each population we built a negative binomial and possion regression the model and then use posterior predictive checks to compare the models.
+We also compare between the models using the resampling method leave-one-out (LOO) cross-validation (CV).
+The results for the models based on each population is as follows:
+
+### White Non-Hispanic Population
+
+Posterior Predictive Check: 
+```{r}
+#| echo: FALSE
+#| message: FALSE
+#| warning: FALSE
+#| label: fig-ppcheckpoissonvsbinomial-white
+#| layout-ncol: 2
+#| fig-cap: "Comparing posterior prediction checks for Poisson and negative binomial models in Leading cause of death of White Non-Hispanic Population in New York City"
+#| fig-subcap: ["Negative binomial model", "Poisson model"]
+
+library(magick)
+
+# Load the images
+image1 <- image_read("../data/analysis_data/Model_Output/pp_check_neg_binom_white.png")
+image2 <- image_read("../data/analysis_data/Model_Output/pp_check_poisson_white.png")
+
+image1
+image2
+```
+
+Leave-one-out (LOO) cross-validation (CV):
+
+| Model                                    | elpd_diff | se_diff |
+|------------------------------------------|-----------|---------|
+| Negative Binomial                        | -43.4     | 4.0     |
+| Poisson                                  | 0.0       | 0.0     |
+
+### Black Non Hispanic Population 
+Posterior Predictive Check: 
+```{r}
+#| echo: FALSE
+#| message: FALSE
+#| warning: FALSE
+#| label: fig-ppcheckpoissonvsbinomial-black
+#| layout-ncol: 2
+#| fig-cap: "Comparing posterior prediction checks for Poisson and negative binomial models in Leading cause of death of Black Non-Hispanic Population "
+#| fig-subcap: ["Negative binomial model", "Poisson model"]
+
+library(magick)
+
+# Load the images
+image1 <- image_read("../data/analysis_data/Model_Output/pp_check_neg_binom_black.png")
+image2 <- image_read("../data/analysis_data/Model_Output/pp_check_poisson_black.png")
+
+image1
+image2
+```
+
+Leave-one-out (LOO) cross-validation (CV):
+
+| Model                                    | elpd_diff | se_diff |
+|------------------------------------------|-----------|---------|
+| Negative Binomial                        | -37.6     | 3.0     |
+| Poisson                                  | 0.0       | 0.0     |
+
+### Hispanic Population
+
+Posterior Predictive Check: 
+```{r}
+#| echo: FALSE
+#| message: FALSE
+#| warning: FALSE
+#| label: fig-ppcheckpoissonvsbinomial-hispanic
+#| layout-ncol: 2
+#| fig-cap: "Comparing posterior prediction checks for Poisson and negative binomial models in Leading cause of death of Hispanic Population in New York City"
+#| fig-subcap: ["Negative binomial model", "Poisson model"]
+
+library(magick)
+
+# Load the images
+image1 <- image_read("../data/analysis_data/Model_Output/pp_check_neg_binom_hispanic.png")
+image2 <- image_read("../data/analysis_data/Model_Output/pp_check_poisson_hispanic.png")
+
+image1
+image2
+```
+
+Leave-one-out (LOO) cross-validation (CV):
+
+| Model                                    | elpd_diff | se_diff |
+|------------------------------------------|-----------|---------|
+| Negative Binomial                        | -34.0     | 2.8     |
+| Poisson                                  | 0.0       | 0.0     |
+
+### Asian And Pacific Islanders Population
+
+Posterior Predictive Check: 
+```{r}
+#| echo: FALSE
+#| message: FALSE
+#| warning: FALSE
+#| label: fig-ppcheckpoissonvsbinomial-asian
+#| layout-ncol: 2
+#| fig-cap: "Comparing posterior prediction checks for Poisson and negative binomial models in Leading cause of death of Asian And Pacific Islanders Population in New York City"
+#| fig-subcap: ["Negative binomial model", "Poisson model"]
+
+library(magick)
+
+# Load the images
+image1 <- image_read("../data/analysis_data/Model_Output/pp_check_neg_binom_asian.png")
+image2 <- image_read("../data/analysis_data/Model_Output/pp_check_poisson_asian.png")
+
+image1
+image2
+```
+
+Leave-one-out (LOO) cross-validation (CV):
+
+| Model                                    | elpd_diff | se_diff |
+|------------------------------------------|-----------|---------|
+| Negative Binomial                        | -25.7     | 1.4     |
+| Poisson                                  | 0.0       | 0.0     |
 

### 2024-03-17 model_white.R
added average age adjusted death rate

@@ -1,49 +1,0 @@
-
-# load libraries
-library(dplyr)
-library(readr)
-library(ggplot2)
-library(forcats)
-library(rstanarm)
-library(knitr)
-library(kableExtra)
-
-# load the dataset 
-file_path <- "../data/analysis_data/clean_nyc_data.csv"
-nyc_data <- read.csv(file_path)
-
-
-# Convert Deaths to numeric if it's not already
-nyc_data$Deaths <- as.numeric(as.character(nyc_data$Deaths))
-
-# Filter for the year 2014 and the specific race
-race_name <- "White Non-Hispanic" 
-
-nyc_data_2014_race <- nyc_data %>%
-  filter(Year == 2014, Race.Ethnicity == race_name)
-
-# Identify the top 5 causes of death for the selected race
-top_causes_by_race <- nyc_data_2014_race %>%
-  group_by(Leading.Cause) %>%
-  summarise(Total_Deaths = sum(as.numeric(Deaths), na.rm = TRUE), .groups = 'drop') %>%
-  arrange(desc(Total_Deaths)) %>%
-  slice_max(Total_Deaths, n = 5)
-
-# Ensure data for regression includes only the top causes
-nyc_data_for_regression <- nyc_data_2014_race %>%
-  inner_join(top_causes_by_race, by = "Leading.Cause")
-
-# negative binomial regression
-neg_binom_model <- stan_glm(
-  Total_Deaths ~ Leading.Cause,
-  data = nyc_data_for_regression,
-  family = neg_binomial_2(link = "log"),
-  seed = 853
-)
-
-# Output the model summary
-model_summary <- summary(neg_binom_model)
-
-# Display the model summary table
-kable(model_summary, caption = paste("Negative Binomial Regression Summary for", race_name)) %>%
-  kable_styling(bootstrap_options = c("striped", "hover"))

### 2024-03-17 paper/paper.qmd
adding new graphs and tables to quarto

--- 2024-03-16 paper/paper.qmd
@@
 date-format: long
-abstract: "We analysed the leading causes of death in New York City, segmented by sex and ethnicity, from 2007 to 2014.
-Utilizing data derived from NYC death certificates, our research offers insights into public health trends, and the importance of understanding how these causes of death vary across race."
+abstract: "Using Morbidity data for New York City, we analyzed the leading causes of death in New York City, segmented by sex and ethnicity, from 2007 to 2014.
+Utilizing data derived from NYC death certificates, our research offers insights into public health trends and the importance of understanding how these causes of death vary across races."
 format: pdf
@@
 
-You can and should cross-reference sections and sub-sections.
-We use @citeR and @rohan.
-
-The remainder of this paper is structured as follows.
-@sec-data....
+The causes of mortality and determinants of such mortality causes have puzzled researchers for centuries.
+Despite evidence of the cause of mortality identification inaccuracies (Johns et al., 2013), comprehensive and detailed death certificate records along with more powerful computing and statistical analysis tools have made it much easier to observe actual causes of mortality – at least in the developed world.
+The determinants of mortality are much harder to pinpoint, with the relative strength of each determinant an even more daunting task.
+A popular theory posits that “the wealthier the healthier” – that is, higher incomes directly cause lower mortality, and that higher incomes are the leading cause of lower mortality.
+Yet this supposed causal relationship can likewise work in reverse in which higher mortality means lower incomes, and studies have shown this explains much of the correlation (Cutler et al., 2006, p.
+117).
+Surprisingly it has also been observed that recessions seem to improve health outcomes, as people are more likely to do less health-damaging activities (i.e., drinking and smoking) and more health-enhancing activities (i.e., exercise) (Ruhm, 2000).
+Another theory suggests that education and ethnicity are the leading causes of lower mortality.
+For example, according to a study, higher mortality is observed among children of black mothers relative to children of white mothers, regardless of the education level of the mother.
+However, when comparing only children of white mothers, mothers with higher education levels are observed to show much lower relative mortality for their children (Pamuk et al., 1998).
+Higher education levels among women are consistently correlated with lower mortality from cardiovascular disease and lung cancer, but also higher mortality from breast cancer (Cutler et al., 2006, p.
+113) – demonstrating the importance of comparing mortality by cause.
+From a long-run perspective, there was hardly any observed difference in life expectancy among England's rich and poor before the Enlightenment period.
+Yet past this period, there is a widespread general divergence.
+Since education has always been more accessible for the rich post-Enlightenment, this highlights the strong potential causal link between education (i.e., the development of new ideas of personal and public health such as germ theory of disease, and openness to adjust behaviours or implement policies validating such ideas) and mortality (Cutler et al., 2006, p.
+117).
+Finally, there is a theory that explains differences in mortality by cause due to differing socio-cultural practices by ethnicity and education levels.
+For example, lung cancer and cardiovascular diseases should be lower among highly educated people since they are less likely to smoke (Pamuk et al., 1998).
+While this theory has its merits, the size of its effect is questionable.
+For example, smoking prevalence is higher among white Americans relative to black Americans, yet this cannot explain black-white differences in American mortality (Cutler et al., 2006, p.
+114).
+This overview of mortality causes and the debate on determinants is the motivating context behind this paper.
+We wish to partially close the gap in understanding between mortality causes and its determinants by analyzing recent data on leading causes of death in New York City, taking a deeper dive into mortality figures by cause and ethnic demographics.
+We model the leading causes of death by rank and magnitude by ethnicity, hypothesizing that each ethnicity follows a negative binomial distribution.
+From there, based on relative trends between each ethnicity such as trends in income, education, and health behaviours that supposedly influence mortality, we can infer to a certain extent the validity of existing mortality determinant theories based on the empirical evidence.
+We find that _______________________________________.
+Each section of this paper is organized as follows.
+Section 2 describes the data and some summary statistics.
+Section 3 describes our modelling, model hypothesis, and justification.
+Section 4 presents descriptive statistics and model results.
+Section 5 is a discussion of the results, including weaknesses, next steps, and conclusive remarks.
+Lastly, all analysis was conducted using R and various packages {citeR}
 
@@
 ## Variables and Data Examination
-The dataset comprises several key variables, including: 
-- Year: The year of the recorded death.
+
+The dataset comprises several key variables, including: - Year: The year of the recorded death.
 
@@
 ## Alternative Datasets and Justification
+
 While other mortality databases exist, such as the CDC's National Center for Health Statistics, the chosen dataset offers a detailed and localized perspective specific to New York City.
@@
 ## Summary Statistics and Relationships
+
 Initial exploratory data analysis revealed key trends and disparities in mortality rates across different populations within New York City.
@@
 
-For our Bayesian regression framework, we employ the stan_glm() function from the rstanarm package, which assumes default normal priors for the regression coefficients as follows:
-$\beta_0$ ~ Normal(0,2.5)
-$\beta_1$ ~ Normal(0,2.5)
+For our Bayesian regression framework, we employ the stan_glm() function from the rstanarm package, which assumes default normal priors for the regression coefficients as follows: $\beta_0$ \~ Normal(0,2.5) $\beta_1$ \~ Normal(0,2.5)
 
 ### Model justification
+
 The negative binomial regression was used due to its ability to account for the over-dispersion present in the data.
@@
 ```
+
 \newpage
@@
 ```
+
+\newpage
+
+```{r, echo=FALSE, message=FALSE}
+
+avg_rate_top_causes_by_race <- nyc_data %>%
+  group_by(Race.Ethnicity, Leading.Cause) %>%
+  summarise(Average_Age_Adjusted_Death_Rate = mean(as.numeric(Age_Adjusted_Death_Rate), na.rm = TRUE)) %>%
+  arrange(Race.Ethnicity, desc(Average_Age_Adjusted_Death_Rate)) %>%
+  mutate(rank = row_number()) %>%
+  filter(rank <= 5) %>%
+  ungroup()
+
+# Rename the columns for display
+colnames(avg_rate_top_causes_by_race) <- c("Race/Ethnicity", "Cause of Death", "Average Age-Adjusted Death Rate", "Rank")
+
+# create table 
+knitr::kable(avg_rate_top_causes_by_race, caption = "Average Age-Adjusted Death Rates by Top 5 Leading Causes of Death by Race") %>%
+
+  kable_styling(bootstrap_options = c("striped", "hover"),
+                full_width = F) %>%
+  column_spec(1, bold = TRUE, width = "3cm") %>%
+  column_spec(2, bold = TRUE, color = "blue", width = "3cm") %>%
+  column_spec(3, width = "3cm")
+
+```
+
+```{r, echo=FALSE, message=FALSE, fig.cap="Top 5 Causes of Death by Average Age-Adjusted Death Rate.", fig.width=10, fig.height=8, echo=FALSE}
+
+# Prepare the data
+avg_rate_top_causes_by_race <- nyc_data %>%
+  mutate(Age_Adjusted_Death_Rate = as.numeric(Age_Adjusted_Death_Rate)) %>%
+  group_by(Race.Ethnicity, Leading.Cause) %>%
+  summarise(Average_Age_Adjusted_Death_Rate = mean(Age_Adjusted_Death_Rate, na.rm = TRUE), .groups = 'drop') %>%
+  arrange(Race.Ethnicity, desc(Average_Age_Adjusted_Death_Rate)) %>%
+  mutate(rank = row_number()) %>%
+  filter(rank <= 5) %>%
+  select(Race.Ethnicity, Leading.Cause, Average_Age_Adjusted_Death_Rate)
+
+# Plotting
+gg <- ggplot(avg_rate_top_causes_by_race, aes(x = `Leading.Cause`, y = Average_Age_Adjusted_Death_Rate, fill = `Leading.Cause`)) +
+  geom_col() + 
+  facet_wrap(~`Race.Ethnicity`, scales = "free_y") + 
+  scale_fill_brewer(palette = "Paired") + 
+  theme_minimal() + 
+  theme(axis.text.x = element_text(angle = 40, hjust = 1), 
+        strip.text.x = element_text(size = 6), 
+        legend.position = "none") + 
+  labs(x = "Top 5 Causes of Death", y = "Average Age-Adjusted Death Rate")
+
+print(gg)
+ggsave("avg_rate_top_causes_by_race.png", plot = gg, width = 12, height = 8)
+
+```
+
+## Model Results
+
+For each population we built a negative binomial and possion regression the model and then use posterior predictive checks to compare the models.
+We also compare between the models using the resampling method leave-one-out (LOO) cross-validation (CV).
+The results for the models based on each population is as follows:
+
+### White Non-Hispanic Population
+
+Posterior Predictive Check:
+
+```{r}
+#| echo: FALSE
+#| message: FALSE
+#| warning: FALSE
+#| label: fig-ppcheckpoissonvsbinomial-white
+#| layout-ncol: 2
+#| fig-cap: "Comparing posterior prediction checks for Poisson and negative binomial models in Leading cause of death of White Non-Hispanic Population in New York City"
+#| fig-subcap: ["Negative binomial model", "Poisson model"]
+
+library(magick)
+
+# Load the images
+image1 <- image_read("../data/analysis_data/Model_Output/pp_check_neg_binom_white.png")
+image2 <- image_read("../data/analysis_data/Model_Output/pp_check_poisson_white.png")
+
+image1
+image2
+```
+
+Leave-one-out (LOO) cross-validation (CV):
+
+| Model             | elpd_diff | se_diff |
+|-------------------|-----------|---------|
+| Negative Binomial | -43.4     | 4.0     |
+| Poisson           | 0.0       | 0.0     |
+
+### Black Non Hispanic Population
+
+Posterior Predictive Check:
+
+```{r}
+#| echo: FALSE
+#| message: FALSE
+#| warning: FALSE
+#| label: fig-ppcheckpoissonvsbinomial-black
+#| layout-ncol: 2
+#| fig-cap: "Comparing posterior prediction checks for Poisson and negative binomial models in Leading cause of death of Black Non-Hispanic Population "
+#| fig-subcap: ["Negative binomial model", "Poisson model"]
+
+library(magick)
+
+# Load the images
+image1 <- image_read("../data/analysis_data/Model_Output/pp_check_neg_binom_black.png")
+image2 <- image_read("../data/analysis_data/Model_Output/pp_check_poisson_black.png")
+
+image1
+image2
+```
+
+Leave-one-out (LOO) cross-validation (CV):
+
+| Model             | elpd_diff | se_diff |
+|-------------------|-----------|---------|
+| Negative Binomial | -37.6     | 3.0     |
+| Poisson           | 0.0       | 0.0     |
+
+### Hispanic Population
+
+Posterior Predictive Check:
+
+```{r}
+#| echo: FALSE
+#| message: FALSE
+#| warning: FALSE
+#| label: fig-ppcheckpoissonvsbinomial-hispanic
+#| layout-ncol: 2
+#| fig-cap: "Comparing posterior prediction checks for Poisson and negative binomial models in Leading cause of death of Hispanic Population in New York City"
+#| fig-subcap: ["Negative binomial model", "Poisson model"]
+
+library(magick)
+
+# Load the images
+image1 <- image_read("../data/analysis_data/Model_Output/pp_check_neg_binom_hispanic.png")
+image2 <- image_read("../data/analysis_data/Model_Output/pp_check_poisson_hispanic.png")
+
+image1
+image2
+```
+
+Leave-one-out (LOO) cross-validation (CV):
+
+| Model             | elpd_diff | se_diff |
+|-------------------|-----------|---------|
+| Negative Binomial | -34.0     | 2.8     |
+| Poisson           | 0.0       | 0.0     |
+
+### Asian And Pacific Islanders Population
+
+Posterior Predictive Check:
+
+```{r}
+#| echo: FALSE
+#| message: FALSE
+#| warning: FALSE
+#| label: fig-ppcheckpoissonvsbinomial-asian
+#| layout-ncol: 2
+#| fig-cap: "Comparing posterior prediction checks for Poisson and negative binomial models in Leading cause of death of Asian And Pacific Islanders Population in New York City"
+#| fig-subcap: ["Negative binomial model", "Poisson model"]
+
+library(magick)
+
+# Load the images
+image1 <- image_read("../data/analysis_data/Model_Output/pp_check_neg_binom_asian.png")
+image2 <- image_read("../data/analysis_data/Model_Output/pp_check_poisson_asian.png")
+
+image1
+image2
+```
+
+Leave-one-out (LOO) cross-validation (CV):
+
+| Model             | elpd_diff | se_diff |
+|-------------------|-----------|---------|
+| Negative Binomial | -25.7     | 1.4     |
+| Poisson           | 0.0       | 0.0     |
 
@@
     -   Yes, the dataset has been used for annual health reports and studies on mortality trends within New York City.
--    Is there a repository that links to any or all papers or systems that use the dataset? If so, please provide a link or other access point.
-    -   No.
+-   Is there a repository that links to any or all papers or systems that use the dataset? If so, please provide a link or other access point. - No.
 -   What (other) tasks could the dataset be used for?

### 2024-03-17 scripts/04-causes_of_death_results.R

@@ -27,6 +27,14 @@ death_count_by_cause <- nyc_data %>%
 knitr::kable(death_count_by_cause, caption = "Count of Deaths by Leading Cause")
 
 
+# create table showing age-adjusted death rate by cause (average)
+age_adj_death_rate_by_cause <- nyc_data %>%
+  group_by(Leading.Cause) %>%
+  summarise(Average_Age_Adjusted_Death_Rate = mean(as.numeric(Age_Adjusted_Death_Rate), na.rm = TRUE)) %>%
+  arrange(desc(Average_Age_Adjusted_Death_Rate))
+knitr::kable(age_adj_death_rate_by_cause, caption = "Average Age-Adjusted Death Rates by Leading Cause")
+
+
 # create table showing top 5 leading causes of death by race
 top_causes_by_race <- nyc_data %>%
   group_by(Race.Ethnicity, Leading.Cause) %>%
@@ -42,6 +50,21 @@ table_top_causes_by_race <- top_causes_by_race %>%
 
 knitr::kable(table_top_causes_by_race, caption = "Top 5 Leading Causes of Death by Race")
 
+# create table showing average age-adjusted death rates by top 5 leading causes of death by race
+avg_rate_top_causes_by_race <- nyc_data %>%
+  group_by(Race.Ethnicity, Leading.Cause) %>%
+  summarise(Average_Age_Adjusted_Death_Rate = mean(as.numeric(Age_Adjusted_Death_Rate), na.rm = TRUE)) %>%
+  arrange(Race.Ethnicity, desc(Average_Age_Adjusted_Death_Rate)) %>%
+  mutate(rank = row_number()) %>%
+  filter(rank <= 5) %>%
+  ungroup()
+
+table_avg_rate_top_causes_by_race <- avg_rate_top_causes_by_race %>%
+  select(-rank) %>%
+  arrange(Race.Ethnicity, desc(Average_Age_Adjusted_Death_Rate))
+
+knitr::kable(table_avg_rate_top_causes_by_race, caption = "Average Age-Adjusted Death Rates by Top 5 Leading Cause of Death by Race")
+
 # create bar graph to display leading causes of death by race
 gg <- ggplot(top_causes_by_race, aes(x = Leading.Cause, y = Total_Deaths)) +
   geom_col(aes(fill = Leading.Cause)) + 
@@ -55,6 +78,22 @@ gg <- ggplot(top_causes_by_race, aes(x = Leading.Cause, y = Total_Deaths)) +
 
[…]

### 2024-03-17 paper/paper.qmd
updated models and paper

@@
 The causes of mortality and determinants of such mortality causes have puzzled researchers for centuries.
-Despite evidence of the cause of mortality identification inaccuracies (Johns et al., 2013), comprehensive and detailed death certificate records along with more powerful computing and statistical analysis tools have made it much easier to observe actual causes of mortality – at least in the developed world.
+Despite evidence of the cause of mortality identification inaccuracies {Johns2013}, comprehensive and detailed death certificate records along with more powerful computing and statistical analysis tools have made it much easier to observe actual causes of mortality – at least in the developed world.
 The determinants of mortality are much harder to pinpoint, with the relative strength of each determinant an even more daunting task.
 A popular theory posits that “the wealthier the healthier” – that is, higher incomes directly cause lower mortality, and that higher incomes are the leading cause of lower mortality.
-Yet this supposed causal relationship can likewise work in reverse in which higher mortality means lower incomes, and studies have shown this explains much of the correlation (Cutler et al., 2006, p.
-117).
-Surprisingly it has also been observed that recessions seem to improve health outcomes, as people are more likely to do less health-damaging activities (i.e., drinking and smoking) and more health-enhancing activities (i.e., exercise) (Ruhm, 2000).
+Yet this supposed causal relationship can likewise work in reverse in which higher mortality means lower incomes, and studies have shown this explains much of the correlation {Cutler2006}.
+Surprisingly it has also been observed that recessions seem to improve health outcomes, as people are more likely to do less health-damaging activities (i.e., drinking and smoking) and more health-enhancing activities (i.e., exercise) {Ruhm, 2000}.
 Another theory suggests that education and ethnicity are the leading causes of lower mortality.
 For example, according to a study, higher mortality is observed among children of black mothers relative to children of white mothers, regardless of the education level of the mother.
-However, when comparing only children of white mothers, mothers with higher education levels are observed to show much lower relative mortality for their children (Pamuk et al., 1998).
-Higher education levels among women are consistently correlated with lower mortality from cardiovascular disease and lung cancer, but also higher mortality from breast cancer (Cutler et al., 2006, p.
-113) – demonstrating the importance of comparing mortality by cause.
+However, when comparing only children of white mothers, mothers with higher education levels are observed to show much lower relative mortality for their children {Pamuk1998}.
+Higher education levels among women are consistently correlated with lower mortality from cardiovascular disease and lung cancer, but also higher mortality from breast cancer {Cutler2006} – demonstrating the importance of comparing mortality by cause.
 From a long-run perspective, there was hardly any observed difference in life expectancy among England's rich and poor before the Enlightenment period.
 Yet past this period, there is a widespread general divergence.
-Since education has always been more accessible for the rich post-Enlightenment, this highlights the strong potential causal link between education (i.e., the development of new ideas of personal and public health such as germ theory of disease, and openness to adjust behaviours or implement policies validating such ideas) and mortality (Cutler et al., 2006, p.
-117).
+Since education has always been more accessible for the rich post-Enlightenment, this highlights the strong potential causal link between education (i.e., the development of new ideas of personal and public health such as germ theory of disease, and openness to adjust behaviours or implement policies validating such ideas) and mortality {Cutler2006}.
 Finally, there is a theory that explains differences in mortality by cause due to differing socio-cultural practices by ethnicity and education levels.
-For example, lung cancer and cardiovascular diseases should be lower among highly educated people since they are less likely to smoke (Pamuk et al., 1998).
+For example, lung cancer and cardiovascular diseases should be lower among highly educated people since they are less likely to smoke {Pamuk1998}.
 While this theory has its merits, the size of its effect is questionable.
-For example, smoking prevalence is higher among white Americans relative to black Americans, yet this cannot explain black-white differences in American mortality (Cutler et al., 2006, p.
-114).
+For example, smoking prevalence is higher among white Americans relative to black Americans, yet this cannot explain black-white differences in American mortality {Cutler2006}.
 This overview of mortality causes and the debate on determinants is the motivating context behind this paper.

### 2024-03-17 scripts/model_asianpacific.R

@@ -14,7 +14,7 @@ library(knitr)
 library(kableExtra)
 
 # load the dataset 
-file_path <- "../data/analysis_data/clean_nyc_data.csv"
+file_path <- "./data/analysis_data/clean_nyc_data.csv"
 nyc_data <- read.csv(file_path)
 
 
@@ -53,3 +53,6 @@ model_summary <- summary(neg_binom_model)
 # Display the model summary table
 kable(model_summary, caption = paste("Negative Binomial Regression Summary for", race_name)) %>%
   kable_styling(bootstrap_options = c("striped", "hover"))
+
+
+saveRDS(neg_binom_model, file = "./models/neg_binom_model_asian_pacific_islander_2014.rds")

### 2024-03-17 scripts/model_black.R

@@ -14,7 +14,7 @@ library(knitr)
 library(kableExtra)
 
 # load the dataset 
-file_path <- "../data/analysis_data/clean_nyc_data.csv"
+file_path <- "./data/analysis_data/clean_nyc_data.csv"
 nyc_data <- read.csv(file_path)
 
 
@@ -53,3 +53,5 @@ model_summary <- summary(neg_binom_model)
 # Display the model summary table
 kable(model_summary, caption = paste("Negative Binomial Regression Summary for", race_name)) %>%
   kable_styling(bootstrap_options = c("striped", "hover"))
+
+saveRDS(neg_binom_model, file = "./models/neg_binom_model_black_2014.rds")

### 2024-03-17 scripts/model_hispanic.R

@@ -14,7 +14,7 @@ library(knitr)
 library(kableExtra)
 
 # load the dataset 
-file_path <- "../data/analysis_data/clean_nyc_data.csv"
+file_path <- "./data/analysis_data/clean_nyc_data.csv"
 nyc_data <- read.csv(file_path)
 
 
@@ -53,3 +53,5 @@ model_summary <- summary(neg_binom_model)
 # Display the model summary table
 kable(model_summary, caption = paste("Negative Binomial Regression Summary for", race_name)) %>%
   kable_styling(bootstrap_options = c("striped", "hover"))
+
+saveRDS(neg_binom_model, file = "./models/neg_binom_model_hispanic_2014.rds")

### 2024-03-17 scripts/model_notstated.R

@@ -14,7 +14,7 @@ library(knitr)
 library(kableExtra)
 
 # load the dataset 
-file_path <- "../data/analysis_data/clean_nyc_data.csv"
+file_path <- "./data/analysis_data/clean_nyc_data.csv"
 nyc_data <- read.csv(file_path)
 
 
@@ -53,3 +53,5 @@ model_summary <- summary(neg_binom_model)
 # Display the model summary table
 kable(model_summary, caption = paste("Negative Binomial Regression Summary for", race_name)) %>%
   kable_styling(bootstrap_options = c("striped", "hover"))
+
+saveRDS(neg_binom_model, file = "./models/neg_binom_model_not_stated_2014.rds")

### 2024-03-17 scripts/model_other.R

@@ -14,7 +14,7 @@ library(knitr)
 library(kableExtra)
 
 # load the dataset 
-file_path <- "../data/analysis_data/clean_nyc_data.csv"
+file_path <- "./data/analysis_data/clean_nyc_data.csv"
 nyc_data <- read.csv(file_path)
 
 # Convert Deaths to numeric if it's not already
@@ -52,3 +52,6 @@ model_summary <- summary(neg_binom_model)
 # Display the model summary table
 kable(model_summary, caption = paste("Negative Binomial Regression Summary for", race_name)) %>%
   kable_styling(bootstrap_options = c("striped", "hover"))
+
+saveRDS(neg_binom_model, file = "./models/neg_binom_model_other_2014.rds")
+

### 2024-03-17 scripts/model_overall_causes.R

@@ -15,7 +15,7 @@ library(knitr)
 library(kableExtra)
 
 # load the dataset 
-file_path <- "../data/analysis_data/clean_nyc_data.csv"
+file_path <- "./data/analysis_data/clean_nyc_data.csv"
 nyc_data <- read.csv(file_path)
 
 
@@ -45,4 +45,6 @@ neg_binom_model_overall <- stan_glm(
 )
 
 # Summary model
-summary(neg_binom_model_overall) 
+summary(neg_binom_model_overall)
+
+saveRDS(neg_binom_model, file = "./models/neg_binom_model_overall_causes_2014.rds")

### 2024-03-17 scripts/model_white.R

@@ -15,7 +15,7 @@ library(knitr)
 library(kableExtra)
 
 # load the dataset 
-file_path <- "../data/analysis_data/clean_nyc_data.csv"
+file_path <- "./data/analysis_data/clean_nyc_data.csv"
 nyc_data <- read.csv(file_path)
 
 
@@ -53,3 +53,5 @@ model_summary <- summary(neg_binom_model)
 # Display the model summary table
 kable(model_summary, caption = paste("Negative Binomial Regression Summary for", race_name)) %>%
   kable_styling(bootstrap_options = c("striped", "hover"))
+
+saveRDS(neg_binom_model, file = "./models/neg_binom_model_white_2014.rds")

### 2024-03-17 paper/paper.qmd
updated paper

@@
 
-## First discussion point {#sec-first-point}
-
-If my paper were 10 pages, then should be be at least 2.5 pages.
-The discussion is a chance to show off what you know and what you learnt from all this.
-
-## Second discussion point
-
-## Third discussion point
-
-## Weaknesses and next steps
-
-Weaknesses and next steps should also be included.
+Contrary to expectations, we find that in modelling the rank and magnitude of the leading cause of death for each ethnicity, utilizing posterior predictive checks, the Poisson distribution offers a better fit relative to the negative binomial distribution (Figure 3; Figure 4; Figure 5; Figure 6).
+This disproves our hypothesis in which we expected the negative binomial distribution to be a better fit.
+Results are robust to model cross-validation using LOO for each ethnicity.
+This is interesting in that a negative binomial distribution is often considered a better fit for estimating mortality, in large part due to overdispersion concerns, yet the Poisson model proved to be a better fit here given the data.
 
@@
 # Appendix {.unnumbered}
-
-# Additional data details
 

### 2024-03-17 paper/paper.qmd
updated graphs

@@
 
-# Prepare the data
 avg_rate_top_causes_by_race <- nyc_data %>%
-  mutate(Age_Adjusted_Death_Rate = as.numeric(Age_Adjusted_Death_Rate)) %>%
   group_by(Race.Ethnicity, Leading.Cause) %>%
-  summarise(Average_Age_Adjusted_Death_Rate = mean(Age_Adjusted_Death_Rate, na.rm = TRUE), .groups = 'drop') %>%
+  summarise(Average_Age_Adjusted_Death_Rate = mean(as.numeric(Age_Adjusted_Death_Rate), na.rm = TRUE)) %>%
   arrange(Race.Ethnicity, desc(Average_Age_Adjusted_Death_Rate)) %>%
@@
   filter(rank <= 5) %>%
-  select(Race.Ethnicity, Leading.Cause, Average_Age_Adjusted_Death_Rate)
+  ungroup()
 
-# Plotting
-gg <- ggplot(avg_rate_top_causes_by_race, aes(x = `Leading.Cause`, y = Average_Age_Adjusted_Death_Rate, fill = `Leading.Cause`)) +
-  geom_col() + 
-  facet_wrap(~`Race.Ethnicity`, scales = "free_y") + 
+gg <- ggplot(avg_rate_top_causes_by_race, aes(x = Leading.Cause, y = Average_Age_Adjusted_Death_Rate)) +
+  geom_col(aes(fill = Leading.Cause)) + 
+  facet_wrap(~Race.Ethnicity, scales = "free_y") + 
   scale_fill_brewer(palette = "Paired") + 
@@
 print(gg)
-ggsave("avg_rate_top_causes_by_race.png", plot = gg, width = 12, height = 8)
 

### 2024-03-17 paper/paper.qmd
fixed up figure orientation

@@
   theme_minimal() + 
-  theme(axis.text.x = element_text(angle = 40, hjust = 1), 
+  theme(axis.text.x = element_text(angle = 90, hjust = 1), 
         strip.text.x = element_text(size = 6), 

### 2024-03-17 paper/paper.qmd
Update references in paper

@@
 The causes of mortality and determinants of such mortality causes have puzzled researchers for centuries.
-Despite evidence of the cause of mortality identification inaccuracies {Johns2013}, comprehensive and detailed death certificate records along with more powerful computing and statistical analysis tools have made it much easier to observe actual causes of mortality – at least in the developed world.
+Despite evidence of the cause of mortality identification inaccuracies [@Johns2013], comprehensive and detailed death certificate records along with more powerful computing and statistical analysis tools have made it much easier to observe actual causes of mortality – at least in the developed world.
 The determinants of mortality are much harder to pinpoint, with the relative strength of each determinant an even more daunting task.
 A popular theory posits that “the wealthier the healthier” – that is, higher incomes directly cause lower mortality, and that higher incomes are the leading cause of lower mortality.
-Yet this supposed causal relationship can likewise work in reverse in which higher mortality means lower incomes, and studies have shown this explains much of the correlation {Cutler2006}.
-Surprisingly it has also been observed that recessions seem to improve health outcomes, as people are more likely to do less health-damaging activities (i.e., drinking and smoking) and more health-enhancing activities (i.e., exercise) {Ruhm, 2000}.
+Yet this supposed causal relationship can likewise work in reverse in which higher mortality means lower incomes, and studies have shown this explains much of the correlation [@Cutler2006].
+Surprisingly it has also been observed that recessions seem to improve health outcomes, as people are more likely to do less health-damaging activities (i.e., drinking and smoking) and more health-enhancing activities (i.e., exercise) [@Ruhm2000].
 Another theory suggests that education and ethnicity are the leading causes of lower mortality.
 For example, according to a study, higher mortality is observed among children of black mothers relative to children of white mothers, regardless of the education level of the mother.
-However, when comparing only children of white mothers, mothers with higher education levels are observed to show much lower relative mortality for their children {Pamuk1998}.
-Higher education levels among women are consistently correlated with lower mortality from cardiovascular disease and lung cancer, but also higher mortality from breast cancer {Cutler2006} – demonstrating the importance of comparing mortality by cause.
+However, when comparing only children of white mothers, mothers with higher education levels are observed to show much lower relative mortality for their children [@Pamuk1998].
+Higher education levels among women are consistently correlated with lower mortality from cardiovascular disease and lung cancer, but also higher mortality from breast cancer [@Cutler2006] – demonstrating the importance of comparing mortality by cause.
 From a long-run perspective, there was hardly any observed difference in life expectancy among England's rich and poor before the Enlightenment period.
 Yet past this period, there is a widespread general divergence.
-Since education has always been more accessible for the rich post-Enlightenment, this highlights the strong potential causal link between education (i.e., the development of new ideas of personal and public health such as germ theory of disease, and openness to adjust behaviours or implement policies validating such ideas) and mortality {Cutler2006}.
-Finally, there is a theory that explains differences in mortality by cause due to differing socio-cultural practices by ethnicity and education levels.
-For example, lung cancer and cardiovascular diseases should be lower among highly educated people since they are less likely to smoke {Pamuk1998}.
+Since education has always been more accessible for the rich post-Enlightenment, this highlights the strong potential causal link between education (i.e., the development of new ideas of personal and public health such as germ theory of disease, and openness to adjust behaviours or implement policies validating such ideas) and mortality [@Cutler2006].
+Finally, there is a theory that explains differences in mortality by cause due to differing sociology-cultural practices by ethnicity and education levels.
+For example, lung cancer and cardiovascular diseases should be lower among highly educated people since they are less likely to smoke [@Pamuk1998].
 While this theory has its merits, the size of its effect is questionable.
-For example, smoking prevalence is higher among white Americans relative to black Americans, yet this cannot explain black-white differences in American mortality {Cutler2006}.
+For example, smoking prevalence is higher among white Americans relative to black Americans, yet this cannot explain black-white differences in American mortality [@Cutler2006].
 This overview of mortality causes and the debate on determinants is the motivating context behind this paper.
@@
 Section 5 is a discussion of the results, including weaknesses, next steps, and conclusive remarks.
-Lastly, all analysis was conducted using R and various packages {citeR}
+Lastly, all analysis was conducted using R and various packages [@citeR]
 
@@
 
-This analysis utilizes data from the New York City Department of Health and Mental Hygiene, specifically provided by the Bureau of Vital Statistics {@NYCData}.
-The dataset encompasses records of deaths in New York City since 2007, detailing the leading causes of death categorized by sex and ethnicity.
-Each entry in this dataset is derived from NYC death certificates which are the official documentation of every death occurring within the city's jurisdiction.
+This analysis utilizes data from the New York City Department of Health and Mental Hygiene, specifically provided by the Bureau of Vital Statistics [@NYCData].
+The data set encompasses records of deaths in New York City since 2007, detailing the leading causes of death categorized by sex and ethnicity.
+Each entry in this data set is derived from NYC death certificates which are the official documentation of every death occurring within the city's jurisdiction.
 
@@
 
-The dataset comprises several key variables, including: - Year: The year of the recorded death.
+The data set comprises several key variables, including: - Year: The year of the recorded death.
 

### 2024-03-17 paper/paper.qmd
added discussion and fixed abstract

@@
 abstract: "Using Morbidity data for New York City, we analyzed the leading causes of death in New York City, segmented by sex and ethnicity, from 2007 to 2014.
-Utilizing data derived from NYC death certificates, our research offers insights into public health trends and the importance of understanding how these causes of death vary across races."
+Utilizing data derived from NYC death certificates, our research offers insights into public health trends and the importance of understanding how these causes of death vary across races.
+We foud that while some causes of death are consistent across ethnicities, there are significant variations which suggest ethnicity and socioeconomic factors play a crucial role in mortality.
+Lastly, we found that causes of death follow a negative binomial distribution acorss different ethnic groups."
 format: pdf
@@
 The determinants of mortality are much harder to pinpoint, with the relative strength of each determinant an even more daunting task.
+
 A popular theory posits that “the wealthier the healthier” – that is, higher incomes directly cause lower mortality, and that higher incomes are the leading cause of lower mortality.
@@
 For example, smoking prevalence is higher among white Americans relative to black Americans, yet this cannot explain black-white differences in American mortality [@Cutler2006].
+
 This overview of mortality causes and the debate on determinants is the motivating context behind this paper.
@@
 From there, based on relative trends between each ethnicity such as trends in income, education, and health behaviours that supposedly influence mortality, we can infer to a certain extent the validity of existing mortality determinant theories based on the empirical evidence.
-We find that _______________________________________.
+When we fit the leading causes of mortality by ethnicity, including relative magnitudes, we find that Poisson distributions offer a better fit compared to negative binomial distributions.
+When looking at average age-adjusted mortality rates by ethnicity, we find that both education and behavioral differences between ethnicities both significantly influence inter-ethnic differences in mortality rate by cause of death.
+
 Each section of this paper is organized as follows.
@@
 # Results
+
+Results indicate that Poisson distributions proved to be a better fit for mortality incidence for all ethnicities relative to negative binomial distributions.
+The next sections delves into this in greater detail.
+Our descriptive analysis shows diseases of the heart to be the leading cause of death, with causes of death differing between ethnicities by mortality incidence and average age-adjusted mortality rates.
 
@@
 
+Moving on to the descriptive data, Table 1 indicates that heart diseases are the leading cause of death in NYC.
+In Figure 1, we observe a similar cause of death ranking for the top five causes for both Black non-Hispanics and White non-Hispanics.
+Diabetes incidence however outranks influenza and pneumonia incidence for Black non-Hispanics, while chronic lower respiratory diseases seem to have much greater relative incidence among White non-Hispanics.
+
+Average age-adjusted death rates (per 100,000 people) are however a better metric for understanding mortality causes, given that it adjusts for both population and age.
+From Table 2 we can observe that heart diseases still rank the highest, followed by malignant neoplasms and all other causes.
+Influenza and pneumonia rank fourth followed by assault, diabetes, and cerebrovascular disease, while chronic lower respiratory diseases rank ninth.
+
+In Figure 2, we look at average age-adjusted death rates by ethnicity for leading mortality causes.
+We see that Black non-Hispanics have a significantly higher rate of heart disease relative to White non-Hispanics.
+As mentioned in the introduction, higher education levels in women are associated with lower rates of cardiovascular disease and lung cancer.
+We observe here that heart diseases, which are a major component of cardiovascular diseases and are wholly considered cardiovascular diseases, inflict much higher relative mortality rates in Black non-Hispanics – possibly because of lower average education levels relative to White non-Hispanics.
+However, chronic lower respiratory diseases, of which lung cancer is likely part, inflict higher relative mortality rates on White non-Hispanics.
+This means there must be some other factors, such as income or behavioural factors, that is influencing this mortality rate.
+We talked about how White non-Hispanics tend to smoke more than Black non-Hispanics, which may be influencing these findings.
+Meanwhile, Asians and Pacific Islanders have lower rates across the board for all causes of death relative to other ethnicities, except for cerebrovascular disease.
+
+Overall, these results are important for both how to model causes of death and helping update previous theories of mortality determinants by cause.
+Indeed, it seems that education and behavioural factors play a strong role in influencing the cause of death.
+These results are important for helping mitigate inter-racial differences in cause of death, and finding solutions to lowering rates based on improving on determinant factors for each ethnicity.
+Weaknesses include the lack of inclusion of all-cause age-adjusted mortality by ethnicity, which would be interesting to include for making further comparisons.
+We also relied heavily on descriptive data to make these inferences, which cannot be controlled for a myriad of other factors.
+For that regression analyses can help provide better accuracy.
+The next steps would be to improve on internal and external validity of the findings.
+
 \newpage

### 2024-03-17 paper/paper.qmd
final update

@@
 Utilizing data derived from NYC death certificates, our research offers insights into public health trends and the importance of understanding how these causes of death vary across races.
-We foud that while some causes of death are consistent across ethnicities, there are significant variations which suggest ethnicity and socioeconomic factors play a crucial role in mortality.
-Lastly, we found that causes of death follow a negative binomial distribution acorss different ethnic groups."
+We find that a Poisson distribution is a better fit for modelling leading causes of death by ethnicity, in both rank and magnitude.
+We also observe that differences in education and behaviour (i.e., smoking prevalence) among ethnicities likely significantly influence differences in age-adjusted mortality rates by cause of death between ethnicities."
 format: pdf
@@
 
-# Model details
-
-## Posterior predictive check
-
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check.
-This shows...
-
 # Datasheet
@@
 
-## Diagnostics
-
-@fig-stanareyouokay-1 is a trace plot.
-It shows...
-This suggests...
-
-@fig-stanareyouokay-2 is a Rhat plot.
-It shows...
-This suggests...
-
 \newpage
